### 2015-08-09 manuscript/abstract.tex
Paper files.

\begin{abstract}
This paper proposes an application that collects sensor data from smartwatches and smartphones in order to predict blood alcohol content (BAC) or drunkenness in real-time, discreetly, and non-invasively through a machine learning approach. This system could prevent drunk driving or other dangers related to the consumption of alcohol by giving users a way to measure personal BAC without the use of breathalyzers or other typically undesired methods. Using data collected by a custom application from many volunteers, we trained a machine learning model that will work with an Android application to predict the user’s BAC.
\end{abstract}

\begin{IEEEkeywords}
	\lowercase{Sensor systems and applications, Temperature sensors, Machine learning, Prediction methods, Prediction algorithms, Internet of Things, Middleware, Wireless sensor networks}
\end{IEEEkeywords}

### 2015-08-09 manuscript/intro.tex
Paper files.

\section{Introduction}
%%Describe the problem, state your contributions and the organization of the paper.T

Internet of Things (IoT) is a domain that represents the next most exciting technological revolution since the Internet. IoT will bring endless opportunities and impact every corner of our planet. In the healthcare domain, IoT promises to bring personalized health tracking and monitoring ever closer to the consumers. This phenomena is evidenced in a recent Wall Street Journal (June, 29, 2015) article entitled "Staying Connected is Crucial to Staying Healthy". Modern smartphones and related devices now contain more sensors than ever before. Data from sensors can be collected more easily and more accurately. In 2014, it is estimated that 46 million people are using IoT-based health and fitness applications. Currently, the predominant IoT-based health applications are in sports and fitness. However, disease management or preventive care health applications are becoming more prevalent. For example, the real-time preventive care applications such as those for detecting fall in elderly patients is one of the active research areas due to the aging population \cite{Tacconi:2011}. This paper describes a new category of real-time preventive health application that aims to reduce risky behaviors due to intoxication by predicting blood alcohol content (BAC) non-invasively and in real-time.

The Centers for Disease Control and Prevention (CDC) reported \cite{Schoenborn:2013} that between the years 2008 and 2010 roughly two-thirds of adults were current drinkers, but adults between the ages 18 and 24 were most commonly associated with consumption of more than 5 drinks in one day. There is also a known causal relationship between alcohol consumption and risky behaviors \cite{Assaad:2006}. One of the worst risky behaviors from drinking is drunk driving which is a persistent problem worldwide. Innocent people are harmed/killed in accidents where the driver was under the influence. It is reported by the Bureau of Transportation Statistics \cite{Chambers:2012} that in 2010 47.2\% of pedestrian fatalities and 39.9\% of vehicle occupant fatalities are caused by drunk driving.

Being able to calculate BAC is important for the portion of the population that consumes alcohol. Drinkers often find themselves in situations in which they are drinking but also need to plan on driving home. The issue comes when the drinker is not sure if they are sober enough, as in their BAC is below the legal limit, to drive home safely. In many cases, drinkers misjudge their intoxication, believing they will be able to safely get home, when actually their BAC is higher than the legal limit. While there are some ways available for people to measure their BAC, these methods are often invasive, inaccurate, or impractical.

The most commonly known method of measuring BAC is the use of a breathalyzer. These devices are portable, easily accessible, and accurate. However, breathalyzers have a stigma of DUI arrest, fines, and license suspension. Breathalyzers are also non-discreet and invasive because the device requires the user to blow into the device in order to measure BAC. Due to these reasons, people often do not want to use such a device. A noninvasive, discreet, and real-time method of measuring BAC is desired.

In this paper, we propose an android-based IoT middleware architecture and application for collecting sensor data on smartwatches and smartphones to predict BAC in real-time. A study conducted in Washington University \cite{Ward:2015} shows that electronic intervention programs are more successful at reducing college student drinking than general alcohol awareness program. By wearing a smartwatch, a user can be alerted in real time when the BAC is above certain level and programmed with different options of intervention when such an event occurs. For example, the application can send a text message to a trusted family member or friend or can lock the ignition of the car engine of the drinker. Currently, there are no known applications that leverages data from common wearable devices to predict BAC non-invasively and in real time.

The main contributions of the paper are:
\begin{itemize}
	\item A middleware platform that can collect data from any type of device (virtual or physical) that can communicate with android phones. 
	\item A BAC prediction model based on live smartwatch sensor data that enables users to get BAC measurement in real time.
	\item Geographical visualization of BAC predictions.
	\item Collection of a labeled BAC data set for training the prediction model.
\end{itemize}

%****Summary of the results and conclusion****

### 2015-08-09 manuscript/related.tex
Paper files.

\section{Related Work}
%
%Include the current IoT middleware systems that you have surveyed. Should start this early.
%\input{related_work}

There is much research being done in the prediction or calculation of blood alcohol content.

The most common way to calculate BAC is through the Widmark equation. Posey and Mozayani in their paper \cite{Posey:2007} discuss the Widmark equation and improvements on its use by observing 5 previously published models and one new model for BAC prediction. The Widmark equation predicts BAC using elimination rates and the Widmark factor. Current use of the Widmark equation and BAC prediction is discouraged by experts due to inaccuracies, usually overestimation. Mathematical calculation of BAC is still sometimes used in DUI court cases, as there are not many alternatives, but it requires lab test results, information about what the person was drinking, how much and how long, BMI, and other hard to find details such as elimination rates. Some of this information has to be estimated or generalized which causes the calculation to be inaccurate. Also, due to the amount of information required, such a method is unusable by the general public for personal use. While BAC prediction is generally frowned upon in court cases, Posey and Mozayani make a point that using several different models provides a more accurate prediction than using the original Widmark equation alone.

Another known method of measuring BAC is through a transdermal ethanol sensor. Webster and Gabler \cite{Webster:2007} discussed the feasibility of transdermal ethanol sensors being integrated into ignition interlock systems in order to prevent drunk driving. A way to reduce the number of incidents related to drunk driving is to somehow stop people from driving drunk. Their approach to do this is to fit every vehicle with a ignition interlock system that responds to the driver's BAC. Current interlock systems like this involve a breathalyzer, but these are seen as expensive, bulky, and elicit a negative response from the general public. Transdermal ethanol sensing provides a more discreet way to measure BAC. However, according to the paper, an issue with transdermal ethanol sensing is that there is a significant delay, up to roughly an hour, from the time of alcohol consumption and detection at the skin.

In 2014, James Baldwin applied for a patent on a wearable blood alcohol measuring device \cite{Baldwin:2014} as a non-invasive, discreet, and wearable method of continuously predicting blood alcohol content. Baldwin's method involves the use of a transdermal ethanol sensor and a method of predicting future BAC given the user's intended consumption. It is also meant to provide guidance for the user as to when to stop in order to be at a certain BAC by a previously specified time. Baldwin's invention is very similar to this project with the only major difference being that it uses a wearable transdermal ethanol sensor to directly measure BAC instead of using common sensors found in smartwatches and a prediction model, such as we do, in order to predict BAC. The use of a transdermal ethanol sensor is proven to have many issues, mostly in the delay between consumption and sensing \cite{Webster:2007}. Another issue is that it requires a special sensor that would only be used for this purpose. Our project improves on this issue by using a common device that many already own for various other reasons. Also, due to our approach for monitoring drunkenness, our method may provide more immediate predictions.

In 2015, Koukiou and Anastassopoulos published a paper on using a neural network to identify drunkenness from thermal infrared imagery \cite{Koukiou:2015}. Different neural networks were trained on different parts of the face in order to determine which areas can be used to discriminated "a drunk from a sober person." Their findings showed that the forehead was the main feature to change thermal behavior from the consumption of alcohol. Later, the whole face was used to train a neural structure which showed high performance on discrimination. The results showed that the forehead and the nose are the areas of the face that were most prominent in drunk people. A neural networks was then trained on the forehead of only one person which was then used on other persons in order to determine those who were drunk. The results had a success rate of around 90\%. This study shows that there is a correlation between skin temperature and drunkenness, which is applicable to our research because of the skin temperature sensor found on many smartwatches. This project has many similar applications as our would, such as being able to be incorporated into ignition interlock systems or to give evidence to authorities about intoxication in investigations. However, this system may not be able to be used by people to determine their own drunkenness as it requires a thermal infrared camera to be focused on the users face in order to do so. This would be difficult to do discreetly or conveniently.

There is also research going into using various sensors for health applications, many using a machine learning approach to do so.

In "Multi-Sensor Fusion for Enhanced Contextual Awareness of Everyday Activites with Ubiquitous Devices" \cite{Guiry:2014}, the authors attempt to use the sensors in smartphones and smartwatches to identify various daily activities such as walking, running, cycling, sitting, etc. It also identified whether or not the user is indoors or outdoors. Collected data was then used to train 5 different machine learning algorithms: C4.5, CART, Na\"{\i}ve Bayes, Multi-Layer Perceptrons, and Support Vector Machines. Results showed promise, with one model for inside/outside classification being 100\% correct. This study shows that the use of multiple sensors improves classification accuracy. For our research, we may be able to improve BAC prediction by recognizing when the user is taking a drink. This research shows that machine learning models can be used to label collected sensor data to certain activities.

Study has been done in using accelerometers to detect seizures \cite{Milosevic:2014} and using smartphones and smartwatches to monitor Parkinson Disease \cite{Sharma:2014}. In research on the detection of seizures \cite{Milosevic:2014}, researches use accelerometers on each wrist and ankle and a machine learning model in order to detect epileptic convulsions. Results included detection of 19 out of 23 seizures that were longer than 30 seconds each, with only a false alarm rate of .39 per hour. SPARK, a proposed framework that uses smartphones and smartwatches to monitor for symptoms of Parkinson Disease \cite{Sharma:2014}, including speech analysis, facial tremors, dyskinesia, freezing of gait. SPARK is for users to have their body motions monitored for dyskinesia and freezing gait while also having the user read paragraphs for speech analysis or pointing the camera for facial analysis. SPARK will also provide recommendations to physicians based on the collected information, such as changes in medication. SPARK is still being tested in order to determine the usefulness of such a framework for those who would use it. It is noted that there are limitations to it in that many patients will not be able to adopt the technologies required for it. There are also many issues, such as sensor misplacement and log errors, that can be encountered. Both these project show a desire to create health-based monitoring applications.

%Current list of applications on smart watch, and their novelty. Current Prediction of BAC and associated ML algorithms

### 2015-08-09 manuscript/system.tex
Paper files.

\section{System Architecture}
A custom Android application (Figure \ref{app}) was designed to be used during the study. This application collected and locally stored sensor data from a Microsoft Band smartwatch in a CSV file. The following data was collected from the Microsoft Band:
\begin{itemize}
	\item Accelerometer
	\item Gyroscope
	\item Pace
	\item Speed
	\item Distance traveled
	\item Motion type (idle, walking, etc.)
	\item Heart rate
	\item Pedometer
	\item Skin temperature
	\item Ultraviolet level
	\item Contact (if smartwatch is on)
	\item Calories burned
\end{itemize}
Additional data continuously collected and stored in the CSV by the application included:
\begin{itemize}
	\item GPS coordinates from the smartphone 
	\item Button that was pressed whenever the user took a drink
	\item Input field used to periodically record the BAC measured by a separate breathalyzer.
\end{itemize}

\begin{figure}[H]
	\centering
	\includegraphics[scale=.20]{smartwatch.png}
	\caption{Capture of the Android application}
	\label{app}
\end{figure}

User information was also recorded and stored through the application before data collection began. Each user was given an unique user id (UUID) and the user profile included age, sex, body mass index, and blood type which was also stored locally as a JSON file.

After the data had been collected, the stored csv file and json file were uploaded to the web server and data inserted into mySQl database by a php script. Data stored on the MySQL database was then displayed on a web client by a geographical map.

\begin{figure}[H]
	\centering
	\includegraphics[scale=.35]{webpage.png}
	\caption{Current data visualization}
	\label{webpage}
\end{figure}

As of now, the web client displays the last sample for each user at the recorded location (Figure \ref{webpage}). Ideally in future systems, many samples will be averaged across regions (Figure \ref{concept}).

\begin{figure}[H]
	\centering
	\includegraphics[scale=.23]{webconcept.png}
	\caption{Ideal data visualization}
	\label{concept}
\end{figure}

The collected data was also used to train a machine learning model which would be implemented in a future Android application to predict user BAC or drunkenness.

\begin{figure}[H]
	\centering
	\includegraphics[scale=.30]{applicationsystem.png}
	\caption{Application system}
	\label{appsystem}
\end{figure}

Representation of this system can be found in Figure \ref{overallsystem}.

\begin{figure}[H]
	\centering
	\includegraphics[scale=.14]{overall_system.png}
	\caption{Overall system}
	\label{overallsystem}
\end{figure}

### 2015-08-16 manuscript/content/abstract.tex
Switch format and other changes

\begin{abstract}
This paper proposes an application that collects sensor data from smartwatches and smartphones in order to predict blood alcohol content (BAC) or drunkenness in real-time, discreetly, and non-invasively through a machine learning approach. This system could prevent drunk driving or other dangers related to the consumption of alcohol by giving users a way to measure personal BAC without the use of intrusive breathalyzers or guess work.  Using data collected by a custom application from many volunteers, we trained a machine learning model that will work with an Android application to predict the user's BAC in real-time.
\end{abstract}

%\begin{IEEEkeywords}
	%\lowercase{Sensor systems and applications, Temperature sensors, Machine learning, Prediction methods, Prediction algorithms, Internet of Things, Middleware, Wireless sensor networks}
%\end{IEEEkeywords}

### 2015-08-16 manuscript/content/intro.tex
Switch format and other changes

--- 2015-08-09 manuscript/intro.tex
@@
 
-The Centers for Disease Control and Prevention (CDC) reported \cite{Schoenborn:2013} that between the years 2008 and 2010 roughly two-thirds of adults were current drinkers, but adults between the ages 18 and 24 were most commonly associated with consumption of more than 5 drinks in one day.
+The Centers for Disease Control and Prevention (CDC) reported \cite{Schoenborn:2013} that between the years 2008 and 2010, roughly two-thirds of adults were current drinkers, but adults between the ages 18 and 24 were most commonly associated with consumption of more than 5 drinks in one day.
 There is also a known causal relationship between alcohol consumption and risky behaviors \cite{Assaad:2006}.
@@
 Innocent people are harmed/killed in accidents where the driver was under the influence.
-It is reported by the Bureau of Transportation Statistics \cite{Chambers:2012} that in 2010 47.2\% of pedestrian fatalities and 39.9\% of vehicle occupant fatalities are caused by drunk driving.
+It is reported by the Bureau of Transportation Statistics \cite{Chambers:2012} that in 2010, 47.2\% of pedestrian fatalities and 39.9\% of vehicle occupant fatalities are caused by drunk driving.
 
@@
 In many cases, drinkers misjudge their intoxication, believing they will be able to safely get home, when actually their BAC is higher than the legal limit.
-While there are some ways available for people to measure their BAC, these methods are often invasive, inaccurate, or impractical.
+%While there are some ways available for people to measure their BAC, these methods are often invasive, inaccurate, or impractical.
 
-The most commonly known method of measuring BAC is the use of a breathalyzer.
+The most commonly known accurate method of measuring BAC is the use of a breathalyzer.
 These devices are portable, easily accessible, and accurate.
-However, breathalyzers have a stigma of DUI arrest, fines, and license suspension.
+However, using breathalyzers have a stigma of DUI arrest, fines, and license suspension.
 Breathalyzers are also non-discreet and invasive because the device requires the user to blow into the device in order to measure BAC.
@@
 In this paper, we propose an android-based IoT middleware architecture and application for collecting sensor data on smartwatches and smartphones to predict BAC in real-time.
-A study conducted in Washington University \cite{Ward:2015} shows that electronic intervention programs are more successful at reducing college student drinking than general alcohol awareness program.
+A study conducted in Washington University \cite{Ward:2015} shows that electronic intervention programs are more successful at reducing college student drinking than a general alcohol awareness program.
 By wearing a smartwatch, a user can be alerted in real time when the BAC is above certain level and programmed with different options of intervention when such an event occurs.
-For example, the application can send a text message to a trusted family member or friend or can lock the ignition of the car engine of the drinker.
+For example, the application can send a text message to a trusted family member or friend, or the software can lock the ignition of the car engine of the drinker.
 Currently, there are no known applications that leverages data from common wearable devices to predict BAC non-invasively and in real time.
@@
 \begin{itemize}
-	\item A middleware platform that can collect data from any type of device (virtual or physical) that can communicate with android phones. 
+	\item A middleware platform that can collect sensor data from any type of device (virtual or physical) that can communicate with android phones. 
 	\item A BAC prediction model based on live smartwatch sensor data that enables users to get BAC measurement in real time.
-	\item Geographical visualization of BAC predictions.
-	\item Collection of a labeled BAC data set for training the prediction model.
+	\item Geographical visualization of collected sensor data for general education on BAC measurement of certain age group.
+	%\item Collection of a labeled BAC data set for training the prediction model.
 \end{itemize}

### 2015-08-16 manuscript/content/method.tex
Switch format and other changes

\section{Methodology}
Research involved people to volunteer in a study that would take roughly two hours in their own home with no monetary compensation, but alcoholic drinks provided free of charge. During the study, volunteers would be measured (height in inches) and weighed (weight in pounds and converted to Kg). This information was recorded and used to determine the amount of alcohol required for the user to reach a BAC of .08 in 1.5 hours by the following formula:

\begin{figure}[H]
	\centering
	\includegraphics[scale=.40]{formula.png}
	\caption{EBAC Formula}
\end{figure}

The volunteer was then given the Microsoft Band to wear on the wrist of their dominant hand and data collection began. Over the course of two hours, the volunteer participated in simple card games while drinking a predetermined number of 1.5oz. of 60 proof liquor shots. While the volunteer took his drink, a researcher would hold the button on the application in order to label and record the event. Also, right before the second shot and every 25 minutes following, the volunteers BAC was measured with the use of a breathalyzer and then recorded in the application. After 1.5 hours, the volunteer ceased drinking but data collection from the smartwatch and breathalyzer continued. At the end of the two hours, BAC was measured one final time and recorded. Data then collection ended and the Microsoft Band was removed. The volunteer was advised to remain safe within their home until sober.

\begin{figure}[H]
	\centering
	\includegraphics[scale=.28]{datacollectionprocess.png}
	\caption{Data collection}
\end{figure}

Collected data was uploaded to a web server and then stored on a mySQL database to be used for visualization of data on a web client. Data was also reviewed for clean up and to find any correlations between the sensor data and measure BAC. This data was then used to train a machine learning model.

### 2015-08-17 manuscript/content/intro.tex
Intro. edits

--- 2015-08-09 manuscript/intro.tex
@@
 \section{Introduction}
-%%Describe the problem, state your contributions and the organization of the paper.T
+%% Domain introduction, problem, contributions, and organization of the paper.
 
@@
 
-The Centers for Disease Control and Prevention (CDC) reported \cite{Schoenborn:2013} that between the years 2008 and 2010 roughly two-thirds of adults were current drinkers, but adults between the ages 18 and 24 were most commonly associated with consumption of more than 5 drinks in one day.
+The Centers for Disease Control and Prevention (CDC) reported that between the years 2008 and 2010, roughly two-thirds of adults were drinkers, but adults between the ages 18 and 24 were most commonly associated with consumption of more than 5 drinks in one day \cite{Schoenborn:2013}.
 There is also a known causal relationship between alcohol consumption and risky behaviors \cite{Assaad:2006}.
-One of the worst risky behaviors from drinking is drunk driving which is a persistent problem worldwide.
-Innocent people are harmed/killed in accidents where the driver was under the influence.
-It is reported by the Bureau of Transportation Statistics \cite{Chambers:2012} that in 2010 47.2\% of pedestrian fatalities and 39.9\% of vehicle occupant fatalities are caused by drunk driving.
+One of the most dangerous, worldwide (citation needed) problems of drinking is drunk driving.
+This problem is not only a hazard to the drunk drivers, but to pedestrians and other drivers.
+It is reported by the Bureau of Transportation Statistics that in 2010, 47.2\% of pedestrian fatalities and 39.9\% of vehicle occupant fatalities are caused by drunk driving \cite{Chambers:2012}.
 
-Being able to calculate BAC is important for the portion of the population that consumes alcohol.
-Drinkers often find themselves in situations in which they are drinking but also need to plan on driving home.
-The issue comes when the drinker is not sure if they are sober enough, as in their BAC is below the legal limit, to drive home safely.
-In many cases, drinkers misjudge their intoxication, believing they will be able to safely get home, when actually their BAC is higher than the legal limit.
-While there are some ways available for people to measure their BAC, these methods are often invasive, inaccurate, or impractical.
-
-The most commonly known method of measuring BAC is the use of a breathalyzer.
-These devices are portable, easily accessible, and accurate.
-However, breathalyzers have a stigma of DUI arrest, fines, and license suspension.
-Breathalyzers are also non-discreet and invasive because the device requires the user to blow into the device in order to measure BAC.
-Due to these reasons, people often do not want to use such a device.
-A noninvasive, discreet, and real-time method of measuring BAC is desired.
-
-In this paper, we propose an android-based IoT middleware architecture and application for collecting sensor data on smartwatches and smartphones to predict BAC in real-time.
-A study conducted in Washington University \cite{Ward:2015} shows that electronic intervention programs are more successful at reducing college student drinking than general alcohol awareness program.
-By wearing a smartwatch, a user can be alerted in real time when the BAC is above certain level and programmed with different options of intervention when such an event occurs.
-For example, the application can send a text message to a trusted family member or friend or can lock the ignition of the car engine of the drinker.
-Currently, there are no known applications that leverages data from common wearable devices to predict BAC non-invasively and in real time.
+Adults often find themselves on outings where they plan on drinking, but also on later driving home.
+At dangerous levels of intoxication, it can be difficult to judge ones own level of intoxication.
+Instead it would be better to get a definitive measurement; of the BAC, or simply a binary response, "drunk" or "not drunk." To do this, one can buy a breathalyzer and blow into it on every outing where drinking is involved.
+This is not a socially beneficial choice.
+The other option is to use a simple smartphone application to manually calculate BAC.
+However these demand a great deal of involvement from the user.
+To be practical, it would be useful to have some sort of non-invasive and accurate monitoring system that will warn its user if they become too intoxicated.
+A study conducted in Washington University shows that electronic intervention programs are more successful at reducing college student drinking than a general alcohol awareness program \cite{Ward:2015}.
+This system can also be used to warn friends and family, or prevent the operation of the user's car.
 
 The main contributions of the paper are:
+
 \begin{itemize}
-	\item A middleware platform that can collect data from any type of device (virtual or physical) that can communicate with android phones. 
-	\item A BAC prediction model based on live smartwatch sensor data that enables users to get BAC measurement in real time.
-	\item Geographical visualization of BAC predictions.
-	\item Collection of a labeled BAC data set for training the prediction model.
+	\item A machine learning model for the prediction of intoxication level from smartwatch sensor data.
+	\item A general Android-based gateway system that can collect data from any type of physical or virtual sensor accessible by the host smartphone.
+	\item An overall, sensor middleware system, including a geographic visualization service, that may be useful for answering many other important research questions.
 \end{itemize}
 
-%****Summary of the results and conclusion****
+%****Summary****

### 2015-08-23 manuscript/content/intro.tex
New code, data, and some paper changes.

@@
 In the healthcare domain, IoT promises to bring personalized health tracking and monitoring ever closer to the consumers.
-This phenomena is evidenced in a recent Wall Street Journal (June, 29, 2015) article entitled "Staying Connected is Crucial to Staying Healthy".
-Modern smartphones and related devices now contain more sensors than ever before.
-Data from sensors can be collected more easily and more accurately.
-In 2014, it is estimated that 46 million people are using IoT-based health and fitness applications.
-Currently, the predominant IoT-based health applications are in sports and fitness.
-However, disease management or preventive care health applications are becoming more prevalent.
-For example, the real-time preventive care applications such as those for detecting fall in elderly patients is one of the active research areas due to the aging population \cite{Tacconi:2011}.
-This paper describes a new category of real-time preventive health application that aims to reduce risky behaviors due to intoxication by predicting blood alcohol content (BAC) non-invasively and in real-time.
+This phenomena is discussed in a recent Wall Street Journal article, "Why Connected Medicine Is Becoming Vital to Health Care" \cite{Landro:2015}.
+Modern smartphones and smartwatches now contain a more diverse collection of sensors than ever before, and people are warming up to them.
+In January 2014, approximately 46 million US smartphone owners were reported to have used health and fitness applications \cite{Nielsen:2014}.
+Currently, sports and fitness are the predominant foci of IoT-based health applications.
+However, applications in disease management and health care are becoming increasingly prevalent.
+For example, detecting falling of elderly patients \cite{Tacconi:2011}.
 
-The Centers for Disease Control and Prevention (CDC) reported that between the years 2008 and 2010, roughly two-thirds of adults were drinkers, but adults between the ages 18 and 24 were most commonly associated with consumption of more than 5 drinks in one day \cite{Schoenborn:2013}.
-There is also a known causal relationship between alcohol consumption and risky behaviors \cite{Assaad:2006}.
-One of the most dangerous, worldwide (citation needed) problems of drinking is drunk driving.
-This problem is not only a hazard to the drunk drivers, but to pedestrians and other drivers.
-It is reported by the Bureau of Transportation Statistics that in 2010, 47.2\% of pedestrian fatalities and 39.9\% of vehicle occupant fatalities are caused by drunk driving \cite{Chambers:2012}.
+Drunk driving is a dangerous, worldwide problem.
+This problem is not only a hazard to the drunk drivers, but also to pedestrians and other drivers.
+It is reported by the Bureau of Transportation Statistics that in 2010, 47.2\% of pedestrian fatalities and 39.9\% of vehicle occupant fatalities were caused by drunk driving \cite{Chambers:2012}.
+The Centers for Disease Control and Prevention (CDC) reported that between the years 2008 and 2010, roughly two-thirds of adults were drinkers, with adults between the ages 18 and 24 having the greatest association with heavy drinking \cite{Schoenborn:2013}.
 
-Adults often find themselves on outings where they plan on drinking, but also on later driving home.
-At dangerous levels of intoxication, it can be difficult to judge ones own level of intoxication.
-Instead it would be better to get a definitive measurement; of the BAC, or simply a binary response, "drunk" or "not drunk." To do this, one can buy a breathalyzer and blow into it on every outing where drinking is involved.
-This is not a socially beneficial choice.
-The other option is to use a simple smartphone application to manually calculate BAC.
-However these demand a great deal of involvement from the user.
+At dangerous levels of intoxication, it can be difficult to judge ones own drunkenness.
+Instead it would be better to get a definitive measurement of the BAC, or simply a binary response: "drunk" or "not drunk." Compact breathalyzers are probably the best option at the moment, but these are not discreet and require deliberate action by the user.
+The other option is to use a smartphone application to manually calculate BAC, but these demand a greater deal of involvement from the user.
 To be practical, it would be useful to have some sort of non-invasive and accurate monitoring system that will warn its user if they become too intoxicated.
-A study conducted in Washington University shows that electronic intervention programs are more successful at reducing college student drinking than a general alcohol awareness program \cite{Ward:2015}.
+It has been shown that electronic intervention programs are more successful at reducing college student drinking than a general alcohol awareness program \cite{Ward:2015}.
 This system can also be used to warn friends and family, or prevent the operation of the user's car.
 
-The main contributions of the paper are:
-
-\begin{itemize}
+The main contributions of our paper are: \begin{itemize}
 	\item A machine learning model for the prediction of intoxication level from smartwatch sensor data.

### 2015-08-23 manuscript/content/related.tex
Switch format and other changes
New code, data, and some paper changes.

--- 2015-08-09 manuscript/related.tex
@@
 \section{Related Work}
-%
-%Include the current IoT middleware systems that you have surveyed. Should start this early.
-%\input{related_work}
 
-There is much research being done in the prediction or calculation of blood alcohol content.
+There is much research being done in the detection, prediction, or calculation of a person's blood alcohol content.
+The most popular mathematical model seems to be the Widmark equation; as evidenced by its prevalence in internet search results.
+Posey and Mozayani discuss the Widmark equation and improvements on its use by observing 5 previously published models and one new model for BAC prediction (clarify!!
+point?) \cite{Posey:2007}.
+Because of the type of information required, such methods are not accurate for personal use by the general public.
 
-The most common way to calculate BAC is through the Widmark equation.
-Posey and Mozayani in their paper \cite{Posey:2007} discuss the Widmark equation and improvements on its use by observing 5 previously published models and one new model for BAC prediction.
-The Widmark equation predicts BAC using elimination rates and the Widmark factor.
-Current use of the Widmark equation and BAC prediction is discouraged by experts due to inaccuracies, usually overestimation.
-Mathematical calculation of BAC is still sometimes used in DUI court cases, as there are not many alternatives, but it requires lab test results, information about what the person was drinking, how much and how long, BMI, and other hard to find details such as elimination rates.
-Some of this information has to be estimated or generalized which causes the calculation to be inaccurate.
-Also, due to the amount of information required, such a method is unusable by the general public for personal use.
-While BAC prediction is generally frowned upon in court cases, Posey and Mozayani make a point that using several different models provides a more accurate prediction than using the original Widmark equation alone.
-
-Another known method of measuring BAC is through a transdermal ethanol sensor.
-Webster and Gabler \cite{Webster:2007} discussed the feasibility of transdermal ethanol sensors being integrated into ignition interlock systems in order to prevent drunk driving.
-A way to reduce the number of incidents related to drunk driving is to somehow stop people from driving drunk.
-Their approach to do this is to fit every vehicle with a ignition interlock system that responds to the driver's BAC.
-Current interlock systems like this involve a breathalyzer, but these are seen as expensive, bulky, and elicit a negative response from the general public.
-Transdermal ethanol sensing provides a more discreet way to measure BAC.
-However, according to the paper, an issue with transdermal ethanol sensing is that there is a significant delay, up to roughly an hour, from the time of alcohol consumption and detection at the skin.
+Current interlock systems involve a breathalyzer, but these are seen as expensive, bulky, and are not are all discreet.
+Recently, there have been some investigations into using transdermal ethanol sensors to directly measure intoxication level.
+Webster and Gabler took a look at the feasibility of using transdermal ethanol sensors with ignition interlock systems in cars.
+Transdermal sensing provides a more discreet way to measure BAC, but this method is accompanied by the problem of a significant lag between transdermally sensed alcohol concentration and blood alcohol concentration.
+The lag is predictable, but not constant and requires additional information about the drinker \cite{Webster:2007}.
 

### 2015-08-23 bac-prediction/scripts/functions.r

@@ -1,0 +1,45 @@
+# Collection of R functions.
+
+# Scales data to have range from 0 to 1.
+# Author: Mario A. Gutierrez (mag262@txstate.edu)
+# Date: 08/23/2015   Version: 1.00
+#	 x - the data to scale.
+# Returns: the scaled data object.
+rescale_01 <- function(x) {
+  minx <- min(x)
+  result <- (x - minx) / (max(x) - minx)
+  return (result)
+}
+
+# Draws a scatterplot with a curve over it, and a line plot in the background.
+# Author: Mario A. Gutierrez (mag262@txstate.edu)
+# Date: 08/23/2015   Version: 1.02
+#	 data - is the data.frame with the data
+#	 back - the data for the line plot in the back.
+#	 front - the data for the scatterplot in the front.
+#	 use_linear - fit a line to the 'front' data if TRUE, fits a curve otherwise
+# 		(default: TRUE).
+# Returns: the resulting ggplot2 object.
+fadoodle <- function(data, bac, front, front.label="", use_linear=TRUE) {
+	require('ggplot2')
+	
+	ifelse(front.label == "", front.label <- toupper(front), front.label <- toupper(front.label))
+	ylab <- paste("BAC vs.", front.label)
+	
+	result <- ggplot(data, aes_string(x = "timestamp", y = bac)) +
+		geom_line() +
+		geom_point(aes_string(x = "timestamp", y = front), color = "red", size = 1.5) +
+		theme(axis.ticks = element_blank(), axis.text.x = element_blank(), axis.text.y = element_blank()) +
+		ylab(ylab) +
+		xlab("TIME")
+		
+	if (use_linear == FALSE) {
+			result <- result + geom_smooth(aes_string(x = "timestamp", y = front))
+	} else {
+		result <- result + geom_smooth(aes_string(x = "timestamp", y = front), method="lm")
+	}
+	
+	return(result)
+}
+
+

### 2015-08-23 manuscript/content/related.tex
Small change, much more tomorrow.

@@
 
-There is much research being done in the detection, prediction, or calculation of a person's blood alcohol content.
-The most popular mathematical model seems to be the Widmark equation; as evidenced by its prevalence in internet search results.
-Posey and Mozayani discuss the Widmark equation and improvements on its use by observing 5 previously published models and one new model for BAC prediction (clarify!!
-point?) \cite{Posey:2007}.
-Because of the type of information required, such methods are not accurate for personal use by the general public.
+There are a few ways of approaching the problem of determining a person's blood alcohol content.
+One approach is to model mathematically the elimination of ethanol in the human body.
+In this case, the Widmark equation, published in 1932 by E.M.P.
+Widmark, is a very popular one, \begin{equation}
+•C=\frac{A}{rW}-(\beta t)
+\end{equation} where C is the BAC, r and $\beta$ are empirically determined constants, A is the mass of the consumed alcohol, and W is the body weight of the person.
+	These days, there have been several improvements and variants.
+Douglas Posey and Ashraf Mozayani published an excellent article comparing this model using parameters determined by different researchers and discussing different models \cite{Posey:2007}.
+They found that the Widmark equation tends to overestimate, and that there can be significant discrepancy between the results of the different models.
+Despite that, they do provide a rough estimate.
+The problem is that these models also require a good deal of information that prohibit their use in a non-intrusive, drunkenness warning system.
 
-Current interlock systems involve a breathalyzer, but these are seen as expensive, bulky, and are not are all discreet.
-Recently, there have been some investigations into using transdermal ethanol sensors to directly measure intoxication level.
-Webster and Gabler took a look at the feasibility of using transdermal ethanol sensors with ignition interlock systems in cars.
-Transdermal sensing provides a more discreet way to measure BAC, but this method is accompanied by the problem of a significant lag between transdermally sensed alcohol concentration and blood alcohol concentration.
-The lag is predictable, but not constant and requires additional information about the drinker \cite{Webster:2007}.
+Another approach to the problem is simply to directly measure the BAC, such as with the recent investigations into the use of transdermal ethanol sensors.
+Transdermal sensing provides a more discreet way to measure BAC, but this method is accompanied by the problem of a significant time lag between the sensed alcohol concentration and actual blood alcohol concentration.
+Gregory Webster and Hampton Gabler closely investigated this problem.
+They found that the lag is predictable, but variable, and requires additional information about the drinks taken to accurately predict \cite{Webster:2007}.
 

### 2015-08-24 manuscript/content/method.tex
Intro. edits
Paper additions and cleanup.

@@
 \section{Methodology}
-Research involved people to volunteer in a study that would take roughly two hours in their own home with no monetary compensation, but alcoholic drinks provided free of charge.
-During the study, volunteers would be measured (height in inches) and weighed (weight in pounds and converted to Kg).
-This information was recorded and used to determine the amount of alcohol required for the user to reach a BAC of .08 in 1.5 hours by the following formula:
 
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=.40]{formula.png}
-	\caption{EBAC Formula}
-\end{figure}
+In this section, we will describe how we collected the data, we will present an analysis of the data, and then lead into the discussion of the machine learning models used.
 
-The volunteer was then given the Microsoft Band to wear on the wrist of their dominant hand and data collection began.
-Over the course of two hours, the volunteer participated in simple card games while drinking a predetermined number of 1.5oz.
-of 60 proof liquor shots.
-While the volunteer took his drink, a researcher would hold the button on the application in order to label and record the event.
-Also, right before the second shot and every 25 minutes following, the volunteers BAC was measured with the use of a breathalyzer and then recorded in the application.
-After 1.5 hours, the volunteer ceased drinking but data collection from the smartwatch and breathalyzer continued.
-At the end of the two hours, BAC was measured one final time and recorded.
-Data then collection ended and the Microsoft Band was removed.
-The volunteer was advised to remain safe within their home until sober.
+\subsection{Collection}
 
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=.28]{datacollectionprocess.png}
-	\caption{Data collection}
-\end{figure}
+Our collection began with the development of an Android application that connected to a Microsoft Band smartwatch and collected data from all of its available sensors.
+This development led to the general Android-based sensor gateway discussed in the previous section.
+Next, we developed a general procedure for our volunteers to follow during the collection of data.
+Our volunteers were eager to freely contribute the anonymous data used in this paper.
+The system collected the data into a .csv file on the Android smartphone and was also transmitted to a central MySQL server.
 
-Collected data was uploaded to a web server and then stored on a mySQL database to be used for visualization of data on a web client.
-Data was also reviewed for clean up and to find any correlations between the sensor data and measure BAC.
-This data was then used to train a machine learning model.
+The Android platform was version 5.0 (kernel 3.4.0-4432708) running on a Samsung S5 smartphone.
+The Microsoft Band was Build Version 10.2.2818.0 09 R.
+Samples of the sensor data were collected every three seconds, based on the update speed of the Band's heart rate sensor.
+For every sample, the most recent sensor value was used for every sensor if available, else it would use the last updated value; or in the case of the accelerometer and gyroscope sensors, the last three values were averaged with linear weighting (the most recent having the most importance).
+There may be a better weighting, but this weighting was suitable for our purposes.
+
+We designed a simple, two-hour procedure for the collection of the data.
+First, we established some necessary information about the subject to estimate the amount of alcohol necessary to reach 0.08 BAC in a 1.5 hour period using the Widmark equation (1).
+The particular formulation of it we used is the following: \begin{equation}
+SD =  BW \cdot Wt \cdot (EBAC + (MR \cdot DP)) \cdot 0.4690
+\end{equation} where $SD$ is the number of standard drinks (10 grams ethanol), $BW$ is the body water contant (0.58 for men and 0.49 for women), Wt is the body weight in lbs, EBAC is the estimated BAC, MR is the metabolism rate (0.17 for women and 0.18 for men), DP is the drinking period in hours, and  $0.4690 = 0.4536 \div (0.806 \cdot 1.2)$, a combination of two constants from the equation and a converstion from kg to lbs \cite{Andersson:2009}\cite{Wiki:BAC}.
+This amount was used to estimate the number of standard drinks to be consumed over the set time period, distributed over equal intervals.
+During this process, at every 25 minutes we took a measurement of the BAC using a BACtrack Trace\textsuperscript{TM} Pro breathalyzer.
+This measurement interval was determined by the cooldown rate of the breathalyzer.
+The activity chosen for the volunteers to engage in was a card or board game of their choice.
+Drinking stopped before 1.5 hours while collection and BAC measurement continued for another 30-45 minutes.
+
+\subsection{Data Analysis}
+
+\subsection{Feature Selection}
+
+\subsection{Model Building}

### 2015-08-24 manuscript/content/related.tex
Paper additions and cleanup.

@@
 \section{Related Work}
-
 There are a few ways of approaching the problem of determining a person's blood alcohol content.
@@
 •C=\frac{A}{rW}-(\beta t)
-\end{equation} where C is the BAC, r and $\beta$ are empirically determined constants, A is the mass of the consumed alcohol, and W is the body weight of the person.
+\end{equation} where $C$ is the BAC, $r$ and $\beta$ are empirically determined constants, $A$ is the mass of the consumed alcohol, and $W$ is the body weight of the person.
 	These days, there have been several improvements and variants.
@@
 
-Another approach to the problem is simply to directly measure the BAC, such as with the recent investigations into the use of transdermal ethanol sensors.
-Transdermal sensing provides a more discreet way to measure BAC, but this method is accompanied by the problem of a significant time lag between the sensed alcohol concentration and actual blood alcohol concentration.
-Gregory Webster and Hampton Gabler closely investigated this problem.
-They found that the lag is predictable, but variable, and requires additional information about the drinks taken to accurately predict \cite{Webster:2007}.
+Another approach to the problem is simply to measure the BAC directly.
+Transdermal ethanol sensors have been a recent option for this approach.
+These can provide a discreet way to measure intoxication, but they are accompanied by the problem of a significant time lag between the sensed alcohol concentration and actual blood alcohol concentration.
+Gregory D.
+Webster and Hampton C.
+Gabler closely investigated this problem.
+They found that the lag is predictable, but not constant, and requires additional information about the drinks taken to accurately predict \cite{Webster:2007}.
 
-In 2014, James Baldwin applied for a patent on a wearable blood alcohol measuring device \cite{Baldwin:2014} as a non-invasive, discreet, and wearable method of continuously predicting blood alcohol content.
-Baldwin's method involves the use of a transdermal ethanol sensor and a method of predicting future BAC given the user's intended consumption.
-It is also meant to provide guidance for the user as to when to stop in order to be at a certain BAC by a previously specified time.
-Baldwin's invention is very similar to this project with the only major difference being that it uses a wearable transdermal ethanol sensor to directly measure BAC instead of using common sensors found in smartwatches and a prediction model, such as we do, in order to predict BAC.
-The use of a transdermal ethanol sensor is proven to have many issues, mostly in the delay between consumption and sensing \cite{Webster:2007}.
-Another issue is that it requires a special sensor that would only be used for this purpose.
-Our project improves on this issue by using a common device that many already own for various other reasons.
-Also, due to our approach for monitoring drunkenness, our method may provide more immediate predictions.
+Similar to our project, James A.
+Baldwin has a patent on a system involving a wearable transdermal ethanol sensor and a mobile device to capture the information \cite{Baldwin:2014}.
+Baldwin describes his system as using a mathematical model to predict the user's BAC given the transdermal sensor data and information about the drinks the user plans to consume.
+A benefit of our system is that it involves no input by the user about the drinks taken, and the user need not buy a special sensor dedicated to this task alone.
 
-In 2015, Koukiou and Anastassopoulos published a paper on using a neural network to identify drunkenness from thermal infrared imagery \cite{Koukiou:2015}.
-Different neural networks were trained on different parts of the face in order to determine which areas can be used to discriminated "a drunk from a sober person." Their findings showed that the forehead was the main feature to change thermal behavior from the consumption of alcohol.
-Later, the whole face was used to train a neural structure which showed high performance on discrimination.
-The results showed that the forehead and the nose are the areas of the face that were most prominent in drunk people.
-A neural networks was then trained on the forehead of only one person which was then used on other persons in order to determine those who were drunk.
-The results had a success rate of around 90\%.
-This study shows that there is a correlation between skin temperature and drunkenness, which is applicable to our research because of the skin temperature sensor found on many smartwatches.
-This project has many similar applications as our would, such as being able to be incorporated into ignition interlock systems or to give evidence to authorities about intoxication in investigations.
-However, this system may not be able to be used by people to determine their own drunkenness as it requires a thermal infrared camera to be focused on the users face in order to do so.
-This would be difficult to do discreetly or conveniently.
+Aside from measuring BAC directly, or developing a biologically-based mathematical model, machine learning is another good approach.
+Georgia Koukiou and Vassilis Anastassopoulos published research this year in using a neural network to identify drunkenness from thermal infrared images of peoples' faces \cite{Koukiou:2015}.
+Neural networks were trained on different parts of the face in order to determine which areas can be used to classify drunkenness.
+They found the forehead was the most significant facial location to observe for determining the drunkenness of a person.
+Their study takes advantage of the effect of alcohol making blood vessels dilate allowing warm blood to come closer to the skin; which is also an important effect for our research.
+Such a system may be good for ignition interlock systems, or drunk surveillance.
 
-There is also research going into using various sensors for health applications, many using a machine learning approach to do so.
+Outside of BAC studies, there has been plenty of research into detecting other activities using smartphone and smartwatch sensor data.
+In \cite{Guiry:2014}, John J.
+Guiry, Pepijin van de Ven, John Nelson, attempt using the sensor data to identify various daily activities, such as: walking, running, cycling, and sitting.
+In their study, they use several machine learning algorithms for their approach: C4.5, CART, Na\"{\i}ve Bayes, ANN, and SVM.
+Their results showed some promise for better future models, with their model for classifying whether a user is indoors or outdoors being the most impressive.
+Successful models for predicting daily activities will certainly be important in a practical implementation of our system.
+This is because the body's response to alcohol consumption may share significant similarity to exercise, dance, or other activities.
 
-In "Multi-Sensor Fusion for Enhanced Contextual Awareness of Everyday Activites with Ubiquitous Devices" \cite{Guiry:2014}, the authors attempt to use the sensors in smartphones and smartwatches to identify various daily activities such as walking, running, cycling, sitting, etc.
-It also identified whether or not the user is indoors or outdoors.
-Collected data was then used to train 5 different machine learning algorithms: C4.5, CART, Na\"{\i}ve Bayes, Multi-Layer Perceptrons, and Support Vector Machines.
-Results showed promise, with one model for inside/outside classification being 100\% correct.
-This study shows that the use of multiple sensors improves classification accuracy.
-For our research, we may be able to improve BAC prediction by recognizing when the user is taking a drink.
-This research shows that machine learning models can be used to label collected sensor data to certain activities.
-
-Study has been done in using accelerometers to detect seizures \cite{Milosevic:2014} and using smartphones and smartwatches to monitor Parkinson Disease \cite{Sharma:2014}.
-In research on the detection of seizures \cite{Milosevic:2014}, researches use accelerometers on each wrist and ankle and a machine learning model in order to detect epileptic convulsions.
-Results included detection of 19 out of 23 seizures that were longer than 30 seconds each, with only a false alarm rate of .39 per hour.
-SPARK, a proposed framework that uses smartphones and smartwatches to monitor for symptoms of Parkinson Disease \cite{Sharma:2014}, including speech analysis, facial tremors, dyskinesia, freezing of gait.
-SPARK is for users to have their body motions monitored for dyskinesia and freezing gait while also having the user read paragraphs for speech analysis or pointing the camera for facial analysis.
-SPARK will also provide recommendations to physicians based on the collected information, such as changes in medication.
-SPARK is still being tested in order to determine the usefulness of such a framework for those who would use it.
-It is noted that there are limitations to it in that many patients will not be able to adopt the technologies required for it.
-There are also many issues, such as sensor misplacement and log errors, that can be encountered.
-Both these project show a desire to create health-based monitoring applications.
-
-%Current list of applications on smart watch, and their novelty. Current Prediction of BAC and associated ML algorithms
+Using smartphones and smartwatches, there is an active desire to create monitoring applications for serious health problems.
+Such as in \cite{Sharma:2014}, where Vinod Sharma, Kunal Mankodiya, Fernando De La Torre, \textit{et.
+al.}, developed a smartwatch-smartphone system for the monitoring and analysis of data from patients with Parkinson Disease.
+This system, named SPARK, includes the analysis of speech and detection of: facial tremors, dyskinesia, and freezing of gait.
+Their system is intended to provide useful recommendations to physicians based on the collected information.
+They concluded noting some potential problems of a full implementation of their system, the most relevant problem being misplacement of the sensors.
+This may also be a problem for us considering the potential importance of the motion-based accelerometer and gyroscope data.

### 2015-08-24 manuscript/content/system.tex
Switch format and other changes
Paper additions and cleanup.

--- 2015-08-09 manuscript/system.tex
@@
 \section{System Architecture}
-A custom Android application (Figure \ref{app}) was designed to be used during the study.
-This application collected and locally stored sensor data from a Microsoft Band smartwatch in a CSV file.
-The following data was collected from the Microsoft Band:
-\begin{itemize}
-	\item Accelerometer
-	\item Gyroscope
-	\item Pace
-	\item Speed
-	\item Distance traveled
-	\item Motion type (idle, walking, etc.)
-	\item Heart rate
-	\item Pedometer
-	\item Skin temperature
-	\item Ultraviolet level
-	\item Contact (if smartwatch is on)
-	\item Calories burned
-\end{itemize}
-Additional data continuously collected and stored in the CSV by the application included:
-\begin{itemize}
-	\item GPS coordinates from the smartphone 
-	\item Button that was pressed whenever the user took a drink
-	\item Input field used to periodically record the BAC measured by a separate breathalyzer.
-\end{itemize}
 
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=.20]{smartwatch.png}
-	\caption{Capture of the Android application}
-	\label{app}
-\end{figure}
+% Present the overall system architecture here.
 
-User information was also recorded and stored through the application before data collection began.
-Each user was given an unique user id (UUID) and the user profile included age, sex, body mass index, and blood type which was also stored locally as a JSON file.
+\subsection{Android-based Sensor Gateway}
 
-After the data had been collected, the stored csv file and json file were uploaded to the web server and data inserted into mySQl database by a php script.
-Data stored on the MySQL database was then displayed on a web client by a geographical map.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=.35]{webpage.png}
-	\caption{Current data visualization}
-	\label{webpage}
-\end{figure}
-
-As of now, the web client displays the last sample for each user at the recorded location (Figure \ref{webpage}).
-Ideally in future systems, many samples will be averaged across regions (Figure \ref{concept}).
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=.23]{webconcept.png}
-	\caption{Ideal data visualization}
-	\label{concept}
-\end{figure}
-
-The collected data was also used to train a machine learning model which would be implemented in a future Android application to predict user BAC or drunkenness.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=.30]{applicationsystem.png}
-	\caption{Application system}
-	\label{appsystem}
-\end{figure}
-
-Representation of this system can be found in Figure \ref{overallsystem}.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=.14]{overall_system.png}
-	\caption{Overall system}
-	\label{overallsystem}
-\end{figure}
+% Describe the sensor gateway here.

### 2015-08-26 manuscript/content/method.tex
Data analysis section

@@
 
+The Microsoft Band we used has an assortment of interesting physical and virtual sensors that we can look at, including: an accelerometer, a gyroscope, distance, heart rate, pedometer, skin temperature, and ultraviolet level.
+With the sample size of five volunteers and a controlled setting for the experiments, some will not be as useful for this study: distance, pedometer, and ultraviolet level.
+In fact, their usefulness may be limited even with larger datasets.
+
+We begin our analysis with the heart rate (HR) by normalizing the heart rates and BACs per subject.
+This way we can plot and compare the HR values over the BAC and see if there are any obvious patterns.
+Doing this we get the plots shown in Figure \ref{fig:heart_rate_facet}.
+
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{../figs/heart_rates}
+	\caption{Heart Rate and BAC per subject (normalized).}
+	\label{fig:heart_rate_facet}
+\end{figure}
+
+The heart rate increases over time for three of the subjects (0, 2, 3) at different rates, decreases for one (4), and stays level for one (1).
+These observations are consistent with data in \cite{Assaad:2006}; a study where the authors relate low and high HR responses to behavioral traits.
+Interestingly, most of the subjects seem to show two patterns of HR activity.
+One is the baseline HR activity, and another is an excitied HR activity that seems to have better correlation with BAC.
+This is most clearly seen in the plot for subject 0, shown in Figure \ref{fig:heart_rate_split}.
+It's not a consistent pattern, however.
+Subject 1 had a very level heart rate with spikes at the regular drinking intervals, and subject 2 had relatively little variance throughout.
+Overall, it seems that experiencing excitement while intoxicated results in a more exaggerated HR response than while sober.
+This may be useful information for determining drunkenness.
+
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{../figs/heart_rate_split}
+	\caption{Heart rate split into baseline and excited states.}
+	\label{fig:heart_rate_split}
+\end{figure}
+
+Next, we plotted the normalized skin temperature over time on top of the normalized BAC values; this is shown in Figure \ref{fig:skin_temperatures}.
+Visually, we see that the correlation of skin temperature and BAC is highly significant.
+This is because of the well known vasodilation effect of alcohol causing warm blood to come near the surface of the skin \cite{Dekker:1999}.
+It is also known that with a high enough level of intoxication it behaves as a vasoconstrictor and drops skin temperature, which we see with subjects 4 and 5 who had reached the highest BAC of the group.
+Subject 2 had a rapid decrease in temperature at the beginning (we may have started collection too soon before the smartwatch sensor adjusted to his skin temperature), it may be appropriate to discard this portion.
+
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{../figs/skin_temperatures}
+	\caption{Skin Temperature and BAC per subject (normalized).}
+	\label{fig:skin_temperatures}
+\end{figure}
+
+The movement sensors (accelerometer and gyroscope) were not too interesting visually or statistically.
+They might have some hidden patterns that contribute to the performance of a modeling technique.
+We think it's usefulness will be increased if we can use it first to predict whether a user is taking a drink, then provide a time-based aggregate of the estimated number of drinks as a feature in our model to determine drunkenness.
+
 \subsection{Feature Selection}

### 2015-08-26 bac-prediction/scripts/functions.r

@@ -20,7 +20,7 @@ rescale_01 <- function(x) {
 #	 use_linear - fit a line to the 'front' data if TRUE, fits a curve otherwise
 # 		(default: TRUE).
 # Returns: the resulting ggplot2 object.
-fadoodle <- function(data, bac, front, front.label="", use_linear=TRUE) {
+fadoodle <- function(data, bac, front, front.label="", curve='l') {
 	require('ggplot2')
 	
 	ifelse(front.label == "", front.label <- toupper(front), front.label <- toupper(front.label))
@@ -31,15 +31,48 @@ fadoodle <- function(data, bac, front, front.label="", use_linear=TRUE) {
 		geom_point(aes_string(x = "timestamp", y = front), color = "red", size = 1.5) +
 		theme(axis.ticks = element_blank(), axis.text.x = element_blank(), axis.text.y = element_blank()) +
 		ylab(ylab) +
-		xlab("TIME")
-		
-	if (use_linear == FALSE) {
+		xlab("")
+	
+	if (!is.null(curve))	{
+		if (curve == 's') {
 			result <- result + geom_smooth(aes_string(x = "timestamp", y = front))
-	} else {
-		result <- result + geom_smooth(aes_string(x = "timestamp", y = front), method="lm")
+		} else {
+			result <- result + geom_smooth(aes_string(x = "timestamp", y = front), method="lm")
+		}
 	}
+	
+	result <- result + facet_wrap( ~ subject, ncol=3)
 	
 	return(result)
 }
+
+# Draws a scatterplot with a curve over it, and a line plot in the background.
+# Author: Mario A. Gutierrez (mag262@txstate.edu)
+# Date: 08/23/2015   Version: 1.02
+#	 data - is the data.frame with the data
+#	 back - the data for the line plot in the back.
+#	 front - the data for the scatterplot in the front.
+#	 use_linear - fit a line to the 'front' data if TRUE, fits a curve otherwise
+# 		(default: TRUE).
+# Returns: the resulting ggplot2 object.
+splithr <- function(data, bac, front, split_bpm=85, front.label="") {
+	require('ggplot2')
+	
+	ifelse(front.label == "", front.label <- toupper(front), front.label <- toupper(front.label))
+	ylab <- paste("BAC vs.", front.label)
+	
+	colors <- rep("red", nrow(data))
+	colors[data$ms_heart_rate_bpm < split_bpm] <- "blue"
+	
[…]

### 2015-08-28 manuscript/content/method.tex
Methodology section

@@
 
-\subsection{Feature Selection}
+\subsection{Features Setup}
 
-\subsection{Model Building}
+Skin temperature and heart rate will have different ranges of values unique to each subject, so per-subject normalization must be performed to put them on the same scale.
+A simple unity-based normalization (min-max) is used to put the feature values in the range 0 to 1.
+No transformations are applied to the movement data.
+
+\subsection{Models and Tools}
+
+We will be taking a look at BAC prediction as both a classification and regression problem.
+As a classification problem, we will threshold the BAC values at a point where we may want to warn the user.
+This allows the data to be used as a binary classification problem.
+For this we will use: decision trees, kNN, logistic regression, and SVM.
+
+As a regression problem, we will be attempting to predict the observed BAC directly.
+To do this, we will use linear regression, kNN and neural networks.
+Using a regression approach, this will allow a user to select their own threshold rather than the threshold determined in a classification model.
+We may however want the threshold to be fixed.
+
+All work will be performed in R version 3.2.1 using the 'caret' package.
+Specifics of the tuning parameters and setup will be described in the evaluation for each model along with the results.
+The evaluation metrics will include standard classification performance measures, such as: specificity, AUC, and F-score; along with RMSE for the regression models.
+Reported scores will be the mean cross-validation score to get a better idea of model performance on unseen data.

### 2015-09-05 manuscript/content/eval.tex
Model evaluation

\section{Evaluation}

In this section we will take a look at the performance of several different models using the aforementioned features. We will start off with some regression modeling and then move on to some classification modeling. Because of our small number of volunteers, we are unlikely to see great results; though we can at least get an idea of the performance of the models and features we use.

\subsection{Regression}

The first model we will consider a classic and a good starting point: the basic linear model. For these predictions, we split the data 80\% for training and 20\% for testing. The following performance values are the values calculated on the unseen test data. Using only the heart rate data, we found the model acheived extremely low performance, with an R$^2$ of 0.077 and root mean squared error (RMSE) of 0.036. Ideally, we want an R$^2$ value of 1.0 and an RMSE of 0.0. Adding in the skin temperature data boosted the R$^2$ up to 0.268 with 0.032 RMSE. Better, but still bad. Adding in the accelerometer data the values were 0.276 and 0.032, for R$^2$ and RMSE respectively. Even using all our preselected features, the linear model is just not a good fit.

So how does a more flexible model perform? We next trained a neural network model on the data. Training a neural network model is very expensive in terms of time, so we reduced our dataset to 8\% of the original. Of the remaining amount we used 70\% for training the model and 30\% for testing. The neural network itself had two hidden layers of five nodes each. This structure was determined experimentally while optimizing for training time and performance. In the input layer we used the heart rate, skin temperature, and accelerometer features. Figure \ref{fig:nn_all} plots the BAC predictions in of the neural network model on top of the actual values from the test set. The real values are jittered to give a sense of the density of the data, the prediction values in blue are not. Overall, the model achieved a much better R$^2$ value of 0.525, with RMSE of 0.026, than the linear regression model.

\begin{figure}
	\includegraphics[width=1.0\textwidth]{../figs/nn_all}
	\caption{ANN predictions (blue) over actual BAC (black, jitterred) using test data.}
	\label{fig:nn_all}
\end{figure}

\subsection{Classification}

This problem might be better handled as a classification task. To do this we will first have to use a threshold to classify the existing BAC labels as \textit{DRUNK} or \textit{SOBER}. We want to warn a person if they are close to reaching the legal limit of 0.08 BAC. A good time to warn a user is at about 0.065 BAC. This splits the classes into 64\% is drunk, and 36\% sober.

The common logistic regression model will be the first to try for our classification task. To train the model, we split the data set by 80\% for training and 20\% for testing as we did for the linear regression model. We will also go ahead and use the full combination of heart, temperature, and accelerometer features. A logistic regression model outputs values from 0.0 to 1.0, so we need to determine where to best split this output into each class. In Figure \ref{fig:log_pred_density}, we show a plot of the predictions of the model on the test data, using the actual labels to distinguish the output. In this plot, we see that the best threshold value for the logistic model predictions is around 0.32; above that we classify \textit{SOBER} and below that \textit{DRUNK}. Using this split, we achieved a precision score of 0.852, recall of 0.730, and F1-score of 0.786.

\begin{figure}
	\includegraphics[width=1.0\textwidth]{../figs/log_pred_density}
	\caption{Logistic regression output frequency with actual labels.}
	\label{fig:log_pred_density}
\end{figure}

The final model we will consider is the SVM using the Gaussian Radial Basis Function (RBF) kernel. This time we first sampled 50\% of the data and then split this into 80\% for training and 20\% for testing. The initial reduction is performed because the SVM is very time-consuming to train, a bit faster than the ANN. On our 0.065-BAC-thresholded data, we found the SVM acheived great performance with a precision of 0.887, recall of 0.931, and overall F1-score of 0.909 on the test data. Ideally, we would not want to warn our users that they are drunk when they are not actually drunk, so we want to try and optimize the model to be as precise as possible. By modifying the error weighting to train against false positive errors, the SVM model achieved a precision of 0.974, with recall of 0.732, and F1-score of 0.836. Our recall dropped for a higher precision, which is a well-known tradeoff for this kind of tuning, but this is fine. We can set a threshold on our smartphone warning system that if at least a recall fraction of the samples from our smartwatch are classified as \textit{DRUNK}, then there is a precision chance that the user is actually at 0.065 BAC.

### 2015-09-05 bac-prediction/data/subject/combine.r

@@ -1,0 +1,19 @@
+args <- commandArgs(trailingOnly=TRUE)
+
+if (length(args) != 2) {
+  cat("Rscript combine.r #_files output_name")
+} else {
+  ds <- read.csv("00.csv", header=TRUE)
+  ds$subject <- rep(0, nrow(ds))
+
+  cat(nrow(ds), "rows added from 00.csv\n")
+
+  for (i in seq(1, as.numeric(args[1])-1, 1)) {
+    tmp <- read.csv(sprintf("%02d.csv", i), header=TRUE)
+    tmp$subject <- rep(i, nrow(tmp))
+    ds <- rbind(ds, tmp)
+    cat(nrow(tmp), "rows added from", sprintf("%02d.csv", i), "\n")
+  }
+
+  write.csv(ds, sprintf("%s.csv", args[2]), row.names=FALSE)
+}

### 2015-09-05 bac-prediction/scripts/functions.r

@@ -1,62 +1,60 @@
-# Collection of R functions.
+# Collection of R functions by Mario A. Gutierrez.
+# Designed for project DRUNK-SMARTWATCH.
 
 # Scales data to have range from 0 to 1.
 # Author: Mario A. Gutierrez (mag262@txstate.edu)
-# Date: 08/23/2015   Version: 1.00
 #	 x - the data to scale.
 # Returns: the scaled data object.
 rescale_01 <- function(x) {
-  minx <- min(x)
-  result <- (x - minx) / (max(x) - minx)
-  return (result)
+    minx <- min(x)
+    result <- (x - minx) / (max(x) - minx)
+    return (result)
 }
 
 # Draws a scatterplot with a curve over it, and a line plot in the background.
 # Author: Mario A. Gutierrez (mag262@txstate.edu)
-# Date: 08/23/2015   Version: 1.02
 #	 data - is the data.frame with the data
-#	 back - the data for the line plot in the back.
+#	 bac - the data for the BAC line plot in the back.
 #	 front - the data for the scatterplot in the front.
 #	 use_linear - fit a line to the 'front' data if TRUE, fits a curve otherwise
 # 		(default: TRUE).
 # Returns: the resulting ggplot2 object.
 fadoodle <- function(data, bac, front, front.label="", curve='l') {
-	require('ggplot2')
+    require('ggplot2')
 	
-	ifelse(front.label == "", front.label <- toupper(front), front.label <- toupper(front.label))
-	ylab <- paste("BAC vs.", front.label)
+    ifelse(front.label == "", front.label <- toupper(front), front.label <- toupper(front.label))
+    ylab <- paste("BAC vs.", front.label)
 	
-	result <- ggplot(data, aes_string(x = "timestamp", y = bac)) +
-		geom_line() +
-		geom_point(aes_string(x = "timestamp", y = front), color = "red", size = 1.5) +
-		theme(axis.ticks = element_blank(), axis.text.x = element_blank(), axis.text.y = element_blank()) +
-		ylab(ylab) +
-		xlab("")
+    result <- ggplot(data, aes_string(x = "timestamp", y = bac)) +
+        geom_line() +
+        geom_point(aes_string(x = "timestamp", y = front), color = "red", size = 1.5) +
+        theme(axis.ticks = element_blank(), axis.text.x = element_blank(), axis.text.y = element_blank()) +
+        ylab(ylab) +
[…]

### 2015-09-05 manuscript/content/method.tex
Model evaluation
Small change.

@@
 This allows the data to be used as a binary classification problem.
-For this we will use: decision trees, kNN, logistic regression, and SVM.
+For this we will use logistic regression and SVM.
 
 As a regression problem, we will be attempting to predict the observed BAC directly.
-To do this, we will use linear regression, kNN and neural networks.
+To do this, we will use linear regression and artificial neural networks (ANN).
 Using a regression approach, this will allow a user to select their own threshold rather than the threshold determined in a classification model.
@@
 
-All work will be performed in R version 3.2.1 using the 'caret' package.
-Specifics of the tuning parameters and setup will be described in the evaluation for each model along with the results.
-The evaluation metrics will include standard classification performance measures, such as: specificity, AUC, and F-score; along with RMSE for the regression models.
-Reported scores will be the mean cross-validation score to get a better idea of model performance on unseen data.
+All work will be performed in R version 3.2.1 using the additional 'neuralnet' and 'kernlab' packages.
+Any necessary information will be provided in the evaluation section.
+To evaulate the models will use some standard performance measures: precision, recall, and F1-score, for the classification models; RMSE and R$^2$ for the regression models.

### 2015-09-05 manuscript/content/conclusions.tex
Conclusions

\section{Conclusions}

We took a look at the design of a general system that could be used for many applications, whether it be a BAC warning system, or a geographical mapping service that displays the smartwatch data of users on a chloropleth map. Part of this system was a general Android-based gateway that can be used to collect data from a variety of sensors connected to an Android smartphone.

After we collected some data, we spent some time to analyze it and look for interesting patterns that were immediately obvious. We found that skin temperature was a good indicator of drunkenness (in our controlled setting). We also discovered that excited heart rate looks to also be a indicator of intoxication level.

Following the overview analysis, we dove into training two regression and classification models. Achieving good performance as a regression problem was difficult. We found that the problem was much better tackled as a classification problem. This worked better because our classification models could ignore a good deal of the variance in the straight BAC predictions. In the end, we found SVM to perform the best on our data.

There are still many other factors to consider in forming a better models. From this research, we found that the accurate prediction of drunkenness in a real application looks possible in theory. There are still many obstacles surrounding the collection of a larger and better data set. Ideally, we would want a data set from a thousand volunteers in candid situations over several days. The two biggest problems are, how do we label the data with the alcohol levels, and how can we model this enormous amount of data in reasonable time? Also, will we need to first determine the activity of the user, or will the other sensor provide sufficient information? And if the former, how can we consistently tag these activities in the data? The answers to this may be simple, or infeasible. In any case, it would be worth it to try and find out.

### 2015-09-06 manuscript/content/system.tex
Final draft complete.

@@
 
-% Present the overall system architecture here.
+Before we get into the specifics of looking into the data and the prediction of the intoxication level, let's take a quick look at the underlying infrastructure of our collection system.
+A huge end-goal of ours is to have a public system where participating users' phones transmit sensor information to a central database through a SOAP-based web service.
+The service stores this information in the database and another data processing system.
+Other users who are not contributing information can connect to our system to access other related services, like a chloropleth map of the sensor data; BAC prediction data in our case, but the system would not limited to it.
+The particular protocols and setup used should be designed to protect the privacy of the sensor data contributors.
 
-\subsection{Android-based Sensor Gateway}
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{../figs/system}
+	\caption{Example general system for large-scale sensor data collection.}
+	\label{fig:system}
+\end{figure}
 
-% Describe the sensor gateway here.
+\subsection{Smartphone Sensor Gateway}
+
+Each of the smartphones in this system are behaving as a type of sensor gateway.
+In order to collect our data, we built such a system on the Android platform.
+Though we used primarily a Microsoft Band, our system allows the easy addition of any sensor that can be connected to from the Android smartphone.
+In Figure \ref{fig:gateway}, we show an overview of the classes used in this system.
+
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{../figs/gateway}
+	\caption{Gateway implementation for our data collection application.}
+	\label{fig:gateway}
+\end{figure}

### 2015-09-06 manuscript/content/abstract.tex
Revisions

@@
 \begin{abstract}
-This paper proposes an application that collects sensor data from smartwatches and smartphones in order to predict blood alcohol content (BAC) or drunkenness in real-time, discreetly, and non-invasively through a machine learning approach.
-This system could prevent drunk driving or other dangers related to the consumption of alcohol by giving users a way to measure personal BAC without the use of intrusive breathalyzers or guess work.
-Using data collected by a custom application from many volunteers, we trained a machine learning model that will work with an Android application to predict the user's BAC in real-time.
+This paper proposes an application that collects sensor data from a smartwatch in order to predict drunkenness in real-time, discreetly, and non-invasively via a machine learning approach.
+This system could prevent drunk driving or other dangers related to the consumption of alcohol by giving users a way to determine personal intoxication level without the use of intrusive breathalyzers or guess work.
+Using smartwatch data collected from several volunteers, we trained a machine learning model that may work with a smartphone application to predict the user's intoxication level in real-time.
 \end{abstract}

### 2015-09-06 manuscript/content/conclusions.tex
Revisions

@@
 
-We took a look at the design of a general system that could be used for many applications, whether it be a BAC warning system, or a geographical mapping service that displays the smartwatch data of users on a chloropleth map.
+We investigated the design of a general system that could be used for many applications, whether it be a BAC warning system, or a geographical mapping service that displays the smartwatch data of users on a chloropleth map.
 Part of this system was a general Android-based gateway that can be used to collect data from a variety of sensors connected to an Android smartphone.
 
-After we collected some data, we spent some time to analyze it and look for interesting patterns that were immediately obvious.
+We then designed an experiment to collect labeled data consistently from volunteers.
+After this, we analyzed the data to discover if there were any interesting patterns that were immediately obvious.
 We found that skin temperature was a good indicator of drunkenness (in our controlled setting).
-We also discovered that excited heart rate looks to also be a indicator of intoxication level.
+We also discovered that excited heart rate looks to also be a good indicator of intoxication level.
 
-Following the overview analysis, we dove into training two regression and classification models.
+Following the overview analysis, we dove into training some regression and classification models.
 Achieving good performance as a regression problem was difficult.

### 2015-09-06 manuscript/content/eval.tex
Revisions

@@
 
-In this section we will take a look at the performance of several different models using the aforementioned features.
-We will start off with some regression modeling and then move on to some classification modeling.
-Because of our small number of volunteers, we are unlikely to see great results; though we can at least get an idea of the performance of the models and features we use.
-
-\subsection{Regression}
-
-The first model we will consider a classic and a good starting point: the basic linear model.
-For these predictions, we split the data 80\% for training and 20\% for testing.
-The following performance values are the values calculated on the unseen test data.
-Using only the heart rate data, we found the model acheived extremely low performance, with an R$^2$ of 0.077 and root mean squared error (RMSE) of 0.036.
-Ideally, we want an R$^2$ value of 1.0 and an RMSE of 0.0.
-Adding in the skin temperature data boosted the R$^2$ up to 0.268 with 0.032 RMSE.
-Better, but still bad.
-Adding in the accelerometer data the values were 0.276 and 0.032, for R$^2$ and RMSE respectively.
-Even using all our preselected features, the linear model is just not a good fit.
-
-So how does a more flexible model perform?
-We next trained a neural network model on the data.
-Training a neural network model is very expensive in terms of time, so we reduced our dataset to 8\% of the original.
-Of the remaining amount we used 70\% for training the model and 30\% for testing.
-The neural network itself had two hidden layers of five nodes each.
-This structure was determined experimentally while optimizing for training time and performance.
-In the input layer we used the heart rate, skin temperature, and accelerometer features.
-Figure \ref{fig:nn_all} plots the BAC predictions in of the neural network model on top of the actual values from the test set.
-The real values are jittered to give a sense of the density of the data, the prediction values in blue are not.
-Overall, the model achieved a much better R$^2$ value of 0.525, with RMSE of 0.026, than the linear regression model.
-
-\begin{figure}
-	\includegraphics[width=1.0\textwidth]{../figs/nn_all}
-	\caption{ANN predictions (blue) over actual BAC (black, jitterred) using test data.}
-	\label{fig:nn_all}
-\end{figure}
+Overall, our data set contains 233,538 samples that were collected from five volunteers.
+Each sample was collected every three seconds from a Microsoft Band smartwatch by our Android data collection system.
+In this section, we evaluate the performance of a few machine learning models on our dataset.
+We use some standard performance measures for our evaluation: precision, recall, and F1-score, for the classification models; RMSE and R$^2$ for the regression models.
+All reported performance values are determined via 5-fold cross-validation.
 
@@
 
-This problem might be better handled as a classification task.
-To do this we will first have to use a threshold to classify the existing BAC labels as \textit{DRUNK} or \textit{SOBER}.
 We want to warn a person if they are close to reaching the legal limit of 0.08 BAC.
 A good time to warn a user is at about 0.065 BAC.
-This splits the classes into 64\% is drunk, and 36\% sober.
+Using this threshold, the classes are split into 64\% is \textit{DRUNK}, and 36\% \textit{SOBER}.
 
-The common logistic regression model will be the first to try for our classification task.
-To train the model, we split the data set by 80\% for training and 20\% for testing as we did for the linear regression model.
-We will also go ahead and use the full combination of heart, temperature, and accelerometer features.
-A logistic regression model outputs values from 0.0 to 1.0, so we need to determine where to best split this output into each class.
+To get a baseline, we first trained a logistic regression model.
+This model outputs values from 0.0 to 1.0, so we need to determine where to best split this output into each class.
 In Figure \ref{fig:log_pred_density}, we show a plot of the predictions of the model on the test data, using the actual labels to distinguish the output.
 In this plot, we see that the best threshold value for the logistic model predictions is around 0.32; above that we classify \textit{SOBER} and below that \textit{DRUNK}.
-Using this split, we achieved a precision score of 0.852, recall of 0.730, and F1-score of 0.786.
+Using this, we achieved a precision score of 0.852, recall of 0.730, and F1-score of 0.786.
 
@@
 
-The final model we will consider is the SVM using the Gaussian Radial Basis Function (RBF) kernel.
-This time we first sampled 50\% of the data and then split this into 80\% for training and 20\% for testing.
-The initial reduction is performed because the SVM is very time-consuming to train, a bit faster than the ANN.
-On our 0.065-BAC-thresholded data, we found the SVM acheived great performance with a precision of 0.887, recall of 0.931, and overall F1-score of 0.909 on the test data.
+Moving forward, we trained a SVM model using the Gaussian Radial Basis Function (RBF) kernel.
+For time constraints, the dataset was reduced by half using a uniform subsample.
+Even so, we found the SVM was able to acheive great performance with a precision of 0.887, recall of 0.931, and overall F1-score of 0.909.
 Ideally, we would not want to warn our users that they are drunk when they are not actually drunk, so we want to try and optimize the model to be as precise as possible.
@@
 We can set a threshold on our smartphone warning system that if at least a recall fraction of the samples from our smartwatch are classified as \textit{DRUNK}, then there is a precision chance that the user is actually at 0.065 BAC.
+
+\subsection{Regression}
+
+So how does this do as a regression problem?
+We first considered the most basic: a linear regression model.
+This did not perform well at all.
+We next trained a neural network (ANN) model on the data.
+This was even slower than the SVM which forced us to subsample to 8\% of the original data.
+The best performing ANN had a structure with two hidden layers of five nodes each.
+Figure \ref{fig:nn_all} plots the BAC predictions in of the neural network model on top of the actual values from the test set.
+The real values are jittered to give a sense of the density of the test data, the prediction values in blue are not.
+This model acheived a R$^2$ value of 0.525, with RMSE of 0.026.
+Much better than the linear regression model, but not nearly as good as the classification models.
+
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{../figs/nn_all}
+	\caption{ANN predictions (blue) over actual BAC (black, jitterred) using test data.}
+	\label{fig:nn_all}
+\end{figure}

### 2015-09-06 manuscript/content/method.tex
Revisions

@@
 
-All work will be performed in R version 3.2.1 using the additional 'neuralnet' and 'kernlab' packages.
-Any necessary information will be provided in the evaluation section.
-To evaulate the models will use some standard performance measures: precision, recall, and F1-score, for the classification models; RMSE and R$^2$ for the regression models.
+All work will be performed in R version 3.2.1 using the additional 'neuralnet' and 'kernlab' packages. Any necessary additional information will be provided in the evaluation section.

### 2015-09-06 manuscript/content/system.tex
Revisions

@@
 Before we get into the specifics of looking into the data and the prediction of the intoxication level, let's take a quick look at the underlying infrastructure of our collection system.
-A huge end-goal of ours is to have a public system where participating users' phones transmit sensor information to a central database through a SOAP-based web service.
+A huge end-goal of ours is to have a public system where participating users' phones transmit sensor information to a central database through a REST-based web service.
 The service stores this information in the database and another data processing system.

### 2015-09-06 manuscript/content/eval.tex
CV log. and SVM_1

@@
 In Figure \ref{fig:log_pred_density}, we show a plot of the predictions of the model on the test data, using the actual labels to distinguish the output.
-In this plot, we see that the best threshold value for the logistic model predictions is around 0.32; above that we classify \textit{SOBER} and below that \textit{DRUNK}.
-Using this, we achieved a precision score of 0.852, recall of 0.730, and F1-score of 0.786.
-
 \begin{figure}
@@
 	\label{fig:log_pred_density}
-\end{figure}
+\end{figure}In this plot, we see that the best threshold value for the logistic model predictions is around 0.32; above that we classify \textit{SOBER} and below that \textit{DRUNK}.
+Using this, we achieved a precision score of 0.855 $\pm$ 0.002, recall of 0.730 $\pm$ 0.004, and F1-score of 0.787 $\pm$ 0.003.
 
@@
 For time constraints, the dataset was reduced by half using a uniform subsample.
-Even so, we found the SVM was able to acheive great performance with a precision of 0.887, recall of 0.931, and overall F1-score of 0.909.
+Even so, we found the SVM was able to acheive great performance with a precision of 0.886 $\pm$ 0.002, recall of 0.930 $\pm$ 0.002, and overall F1-score of 0.907 $\pm$ 0.001.
 Ideally, we would not want to warn our users that they are drunk when they are not actually drunk, so we want to try and optimize the model to be as precise as possible.
@@
 We next trained a neural network (ANN) model on the data.
-This was even slower than the SVM which forced us to subsample to 8\% of the original data.
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{../figs/nn_all}
+	\caption{ANN predictions (blue) over actual BAC (black, jitterred) using test data.}
+	\label{fig:nn_all}
+\end{figure}This was even slower than the SVM which forced us to subsample to 8\% of the original data.
 The best performing ANN had a structure with two hidden layers of five nodes each.
@@
 Much better than the linear regression model, but not nearly as good as the classification models.
-
-\begin{figure}
-	\includegraphics[width=1.0\textwidth]{../figs/nn_all}
-	\caption{ANN predictions (blue) over actual BAC (black, jitterred) using test data.}
-	\label{fig:nn_all}
-\end{figure}

### 2015-09-06 bac-prediction/scripts/functions_old.r

@@ -1,0 +1,457 @@
+# Collection of R functions by Mario A. Gutierrez.
+# Designed for project DRUNK-SMARTWATCH.
+
+# Scales data to have range from 0 to 1.
+# Author: Mario A. Gutierrez (mag262@txstate.edu)
+#	 x - the data to scale.
+# Returns: the scaled data object.
+rescale_01 <- function(x) {
+    minx <- min(x)
+    result <- (x - minx) / (max(x) - minx)
+    return (result)
+}
+
+# Draws a scatterplot with a curve over it, and a line plot in the background.
+# Author: Mario A. Gutierrez (mag262@txstate.edu)
+#	 data - is the data.frame with the data
+#	 bac - the data for the BAC line plot in the back.
+#	 front - the data for the scatterplot in the front.
+#	 use_linear - fit a line to the 'front' data if TRUE, fits a curve otherwise
+# 		(default: TRUE).
+# Returns: the resulting ggplot2 object.
+fadoodle <- function(data, bac, front, front.label="", curve='l') {
+    require('ggplot2')
+	
+    ifelse(front.label == "", front.label <- toupper(front), front.label <- toupper(front.label))
+    ylab <- paste("BAC vs.", front.label)
+	
+    result <- ggplot(data, aes_string(x = "timestamp", y = bac)) +
+        geom_line() +
+        geom_point(aes_string(x = "timestamp", y = front), color = "red", size = 1.5) +
+        theme(axis.ticks = element_blank(), axis.text.x = element_blank(), axis.text.y = element_blank()) +
+        ylab(ylab) +
+        xlab("")
+	
+    if (!is.null(curve))	{
+        if (curve == 's') {
+            result <- result + geom_smooth(aes_string(x = "timestamp", y = front))
+        } else {
+            result <- result + geom_smooth(aes_string(x = "timestamp", y = front), method="lm")
+        }
+    }
+	
+    result <- result + facet_wrap( ~ subject, ncol=3)
+	
+    return(result)
+}
+
+# Draws the heart rate data split in two colors by a set BPM. Used for one of the
+# figures in the paper. Also draws the BAC line in the background.
+# Author: Mario A. Gutierrez (mag262@txstate.edu)
+#	 data - is the data.frame with the data
+#	 bac - the data for the BAC line plot in the back.
[…]

### 2015-09-06 manuscript/content/eval.tex
SVN CV results.

@@
 \end{figure}In this plot, we see that the best threshold value for the logistic model predictions is around 0.32; above that we classify \textit{SOBER} and below that \textit{DRUNK}.
-Using this, we achieved a precision score of 0.855 $\pm$ 0.002, recall of 0.730 $\pm$ 0.004, and F1-score of 0.787 $\pm$ 0.003.
+Using this, we achieved a precision of 0.855 $\pm$ 0.002, recall of 0.730 $\pm$ 0.004, and F1-score of 0.787 $\pm$ 0.003.
 
@@
 For time constraints, the dataset was reduced by half using a uniform subsample.
-Even so, we found the SVM was able to acheive great performance with a precision of 0.886 $\pm$ 0.002, recall of 0.930 $\pm$ 0.002, and overall F1-score of 0.907 $\pm$ 0.001.
-Ideally, we would not want to warn our users that they are drunk when they are not actually drunk, so we want to try and optimize the model to be as precise as possible.
-By modifying the error weighting to train against false positive errors, the SVM model achieved a precision of 0.974, with recall of 0.732, and F1-score of 0.836.
+Even so, we found the SVM was able to achieve great performance with a precision of 0.886 $\pm$ 0.002, recall of 0.930 $\pm$ 0.002, and overall F1-score of 0.907 $\pm$ 0.001.
+Ideally we do not want to warn our users that they are drunk when they are not actually drunk, so we want to try and optimize the model to be as precise as possible.
+By modifying the error weighting to train against false positive errors, the SVM model achieved a precision of 0.970 $\pm$ 0.002, with recall of 0.729 $\pm$ 0.003, and F1-score of 0.832 $\pm$ 0.002.
 Our recall dropped for a higher precision, which is a well-known tradeoff for this kind of tuning, but this is fine.

### 2015-09-06 bac-prediction/scripts/functions.r
script for experiments with cv.

@@ -97,6 +97,13 @@ trainPartition <- function(data, percent) {
     return(result)
 }
 
+# Uniformly shuffles the samples in the data set.
+# Returns: a numeric vector of the shuffled indices.
+shuffleUniform <- function(data) {
+    set.seed(1234) # A constant seed, so we get the same results every time.
+    return(sample(1:nrow(data), nrow(data)))
+}
+
 # Calculate the R-sq value between two vectors.
 #   y - the actual values.
 #   f - the predicted values.
@@ -116,210 +123,21 @@ resid <- function(data, model) {
 # Returns: the RMSE.
 rmse <- function(y, f) { return(sqrt(sum((y - f)^2) / length(y))) }
 
-# Runs linear regression experiments.
-#   data - the raw unnormalized data.
-#   trainSplit - what fraction to use for training.
-runLinearRegression <- function(data, trainSplit, toFile=NULL) {
-    require('ggplot2')
-    if (!is.null(toFile)) sink(file=toFile)
-    
-    cat("trainSplit:", trainSplit,"\n\n")
-    
-    cat("Normalizing columns per subject...\n\n")
-    data$ms_heart_rate_bpm <- rescalePerSubject(data, "ms_heart_rate_bpm")
-    data$ms_skin_temperature_celsius <- rescalePerSubject(data, "ms_skin_temperature_celsius")
-    cat("Heart Rate Summary:\n")
-    print(summary(data$ms_heart_rate_bpm))
-    cat("Skin Temperature Summary:\n")
-    print(summary(data$ms_skin_temperature_celsius))
-    
-    cat("\nGetting train/test partition vector...\n\n")
-    train <- trainPartition(data, trainSplit)
-    test <- !train
-    
-    cat("\n-----------------------------------------------------------\n")
-    # Do linear analysis on skin temperature only.
-    cat("\nTraining lm() on skin temperature...\n")
-    mdl <- lm(bac_observed ~ ms_heart_rate_bpm, data[train,])
-    print(summary(mdl))
-    
-    cat("\nMaking predictions...\n")
-    data$pred <- predict(mdl, data)
-    cat("k.\n\n")
-    
-    data$x <- rep(0, nrow(data))
-    data[test,]$x <- 1:nrow(data[test,])
-    
-    cat("Performance on test data:\n")
[…]

### 2015-09-06 bac-prediction/scripts/functions.r
functions.r final

@@ -316,7 +316,7 @@ runSVM <- function(data, reduce, k, threshold, soberWeight=1, toFile=NULL) {
 #   toFile - the file to save the output to.
 runANN <- function(data, reduce, k, hidden, toFile=NULL) {
     require('ggplot2')
-    require('neuralnet')
+    require('nnet')
     
     if (!is.null(toFile)) sink(file=toFile)
     
@@ -324,9 +324,11 @@ runANN <- function(data, reduce, k, hidden, toFile=NULL) {
     cat("hidden:", hidden, "\n\n")
     
     # INITIAL DATA SETUP.
-    set.seed(1234)
-    data <- data[sample(1:nrow(data), floor(nrow(data)*reduce)),]
-    cat("Dataset subsampled to",nrow(data),"samples.\n")
+    if (reduce < 1) {
+        set.seed(1234)
+        data <- data[sort(sample(1:nrow(data), floor(nrow(data)*reduce))),]
+        cat("Dataset subsampled to",nrow(data),"samples.\n")
+    }
     
     cat("Shuffling dataset.\n")
     shuf <- shuffleUniform(data)
@@ -347,26 +349,28 @@ runANN <- function(data, reduce, k, hidden, toFile=NULL) {
     
     perf <- data.frame(vrsq=numeric(), vrmse=numeric())
     
+    bestTest.rsq <- 0
+    bestTest <- numeric()
+    
     # TRAIN THE NEURAL NETWORK.
     for (i in seq(1, k, 1)) {
         train <- shuf[-sel] # !psize selection of shuffled indices
         test <- shuf[sel] # psize selection of shuffled indices
         
         cat("Fold:",i,"\n")
-        cat("Training set BAC summary:\n")
-        print(summary(data[train,]$bac_observed))
-        cat("Test set BAC summary:\n")
-        print(summary(data[test,]$bac_observed))
         
-        mdl <- neuralnet(bac_observed ~
+        mdl <- nnet(bac_observed ~
             ms_heart_rate_bpm +
             ms_skin_temperature_celsius +
             ms_accelerometer_x +
             ms_accelerometer_y +
             ms_accelerometer_z,
             data=data[train,],
-            hidden=hidden
-        )
+            size=hidden,
+            linout=TRUE,
+            decay = 5e-4,
+            maxit=200
+        );
         
[…]

### 2015-09-06 manuscript/content/eval.tex
Final data.

@@
 	\includegraphics[width=1.0\textwidth]{../figs/nn_all}
-	\caption{ANN predictions (blue) over actual BAC (black, jitterred) using test data.}
+	\caption{ANN predictions (blue) with actual BAC data (black) on test partition.}
 	\label{fig:nn_all}
-\end{figure}This was even slower than the SVM which forced us to subsample to 8\% of the original data.
-The best performing ANN had a structure with two hidden layers of five nodes each.
-Figure \ref{fig:nn_all} plots the BAC predictions in of the neural network model on top of the actual values from the test set.
-The real values are jittered to give a sense of the density of the test data, the prediction values in blue are not.
-This model acheived a R$^2$ value of 0.525, with RMSE of 0.026.
-Much better than the linear regression model, but not nearly as good as the classification models.
+\end{figure}The best performing ANN had a structure with twenty nodes in one hidden layer.
+This was determined by doubling and then reducing node count to find the best performance.
+Additional layers did not show any improvement (the 'neuralnet' package was used to test multiple layers).
+Figure \ref{fig:nn_all} plots the BAC predictions of the ANN along with the actual values using data from a test partition.
+The ANN model achieved a R$^2$ value of 0.524 $\pm$ 0.015, with RMSE of 0.026 $\pm$ 0.000.
+It performed much better than the linear regression model, but not nearly as good as the classification models.

### 2015-09-06 manuscript/content/method.tex
CV log. and SVM_1
Final data.

@@
 This is most clearly seen in the plot for subject 0, shown in Figure \ref{fig:heart_rate_split}.
-It's not a consistent pattern, however.
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{../figs/heart_rate_split}
+	\caption{Heart rate split into baseline and excited states.}
+	\label{fig:heart_rate_split}
+\end{figure}It's not a consistent pattern, however.
 Subject 1 had a very level heart rate with spikes at the regular drinking intervals, and subject 2 had relatively little variance throughout.
@@
 
-\begin{figure}
-	\includegraphics[width=1.0\textwidth]{../figs/heart_rate_split}
-	\caption{Heart rate split into baseline and excited states.}
-	\label{fig:heart_rate_split}
-\end{figure}
-
-Next, we plotted the normalized skin temperature over time on top of the normalized BAC values; this is shown in Figure \ref{fig:skin_temperatures}.
-Visually, we see that the correlation of skin temperature and BAC is highly significant.
+Next, we plotted the normalized skin temperature over time on top of the normalized BAC values; this is shown in Figure \ref{fig:skin_temperatures}. \begin{figure}
+	\includegraphics[width=1.0\textwidth]{../figs/skin_temperatures}
+	\caption{Skin Temperature and BAC per subject (normalized).}
+	\label{fig:skin_temperatures}
+\end{figure}Visually, we see that the correlation of skin temperature and BAC is highly significant.
 This is because of the well known vasodilation effect of alcohol causing warm blood to come near the surface of the skin \cite{Dekker:1999}.
@@
 Subject 2 had a rapid decrease in temperature at the beginning (we may have started collection too soon before the smartwatch sensor adjusted to his skin temperature), it may be appropriate to discard this portion.
-
-\begin{figure}
-	\includegraphics[width=1.0\textwidth]{../figs/skin_temperatures}
-	\caption{Skin Temperature and BAC per subject (normalized).}
-	\label{fig:skin_temperatures}
-\end{figure}
 
@@
 
-All work will be performed in R version 3.2.1 using the additional 'neuralnet' and 'kernlab' packages. Any necessary additional information will be provided in the evaluation section.
+All work will be performed in R version 3.2.1 using the additional 'nnet' and 'kernlab' packages. Any necessary additional information will be provided in the evaluation section.

### 2015-09-29 manuscript/content/conclusions.tex
Fixed typos.

@@
 
-We investigated the design of a general system that could be used for many applications, whether it be a BAC warning system, or a geographical mapping service that displays the smartwatch data of users on a chloropleth map.
+We investigated the design of a general system that could be used for many applications, whether it be a BAC warning system, or a geographical mapping service that displays the smartwatch data of users on a choropleth map.
 Part of this system was a general Android-based gateway that can be used to collect data from a variety of sensors connected to an Android smartphone.

### 2015-09-29 manuscript/content/system.tex
CV log. and SVM_1
Fixed typos.

@@
 
-Before we get into the specifics of looking into the data and the prediction of the intoxication level, let's take a quick look at the underlying infrastructure of our collection system.
 A huge end-goal of ours is to have a public system where participating users' phones transmit sensor information to a central database through a REST-based web service.
 The service stores this information in the database and another data processing system.
-Other users who are not contributing information can connect to our system to access other related services, like a chloropleth map of the sensor data; BAC prediction data in our case, but the system would not limited to it.
+Other users who are not contributing information can connect to our system to access other related services, like a choropleth map of the sensor data; BAC prediction data in our case, but the system would not be limited to it.
 The particular protocols and setup used should be designed to protect the privacy of the sensor data contributors.

### 2015-10-10 manuscript/content/related.tex
Added proper style files, pages are now numbered in accordance with format.
Small revisions to Introduction and Related Work sections.

@@
 There are a few ways of approaching the problem of determining a person's blood alcohol content.
-One approach is to model mathematically the elimination of ethanol in the human body.
+One approach is to devise a mathematical model of the elimination of ethanol in the human body.
 In this case, the Widmark equation, published in 1932 by E.M.P.
-Widmark, is a very popular one, \begin{equation}
-•C=\frac{A}{rW}-(\beta t)
+Widmark, is a very popular one; \begin{equation}C=\frac{A}{rW}-(\beta t) \enspace ,
 \end{equation} where $C$ is the BAC, $r$ and $\beta$ are empirically determined constants, $A$ is the mass of the consumed alcohol, and $W$ is the body weight of the person.
@@
 Gabler closely investigated this problem.
-They found that the lag is predictable, but not constant, and requires additional information about the drinks taken to accurately predict \cite{Webster:2007}.
+They found that the lag is predictable, but not constant, and requires additional information about the number of drinks taken by the user to accurately predict it \cite{Webster:2007}.
 

### 2015-10-12 manuscript/content/method.tex
Still going.

@@
 
-Our collection began with the development of an Android application that connected to a Microsoft Band smartwatch and collected data from all of its available sensors.
-This development led to the general Android-based sensor gateway discussed in the previous section.
+Our collection began with the development of an Android application that can connect to and store data from available sensors on the Microsoft Band smartwatch.
 Next, we developed a general procedure for our volunteers to follow during the collection of data.
 Our volunteers were eager to freely contribute the anonymous data used in this paper.
-The system collected the data into a .csv file on the Android smartphone and was also transmitted to a central MySQL server.
+The data collected by the system was stored into a .csv file on the Android smartphone and also transmitted to a central database server.
 
-The Android platform was version 5.0 (kernel 3.4.0-4432708) running on a Samsung S5 smartphone.
-The Microsoft Band was Build Version 10.2.2818.0 09 R.
+We used Android platform version 5.0 (kernel 3.4.0-4432708) running on a Samsung S5 smartphone.
+The Microsoft Band had Build Version 10.2.2818.0 09 R.
 Samples of the sensor data were collected every three seconds, based on the update speed of the Band's heart rate sensor.
-For every sample, the most recent sensor value was used for every sensor if available, else it would use the last updated value; or in the case of the accelerometer and gyroscope sensors, the last three values were averaged with linear weighting (the most recent having the most importance).
+Each sample used the most recent sensor value if available, otherwise it would use the last value; or in the case of the accelerometer and gyroscope sensors, the three most recent values were averaged with linear weighting (the most recent having highest importance).
 There may be a better weighting, but this weighting was suitable for our purposes.
-
+% HERE.
 We designed a simple, two-hour procedure for the collection of the data.

### 2015-10-12 manuscript/content/method.tex
Revised Methodology.Collection section.

@@
 There may be a better weighting, but this weighting was suitable for our purposes.
-% HERE.
-We designed a simple, two-hour procedure for the collection of the data.
-First, we established some necessary information about the subject to estimate the amount of alcohol necessary to reach 0.08 BAC in a 1.5 hour period using the Widmark equation (1).
+
+The procedure followed by the volunteers during data collection lasted two-hours.
+First, we established some necessary information about the subject in order to estimate the amount of alcohol necessary to reach 0.08 BAC in a 1.5 hour period using the Widmark equation (1).
 The particular formulation of it we used is the following: \begin{equation}
-SD =  BW \cdot Wt \cdot (EBAC + (MR \cdot DP)) \cdot 0.4690
+SD =  BW \cdot Wt \cdot (EBAC + (MR \cdot DP)) \cdot 0.4690 \enspace ,
 \end{equation} where $SD$ is the number of standard drinks (10 grams ethanol), $BW$ is the body water contant (0.58 for men and 0.49 for women), Wt is the body weight in lbs, EBAC is the estimated BAC, MR is the metabolism rate (0.17 for women and 0.18 for men), DP is the drinking period in hours, and  $0.4690 = 0.4536 \div (0.806 \cdot 1.2)$, a combination of two constants from the equation and a converstion from kg to lbs \cite{Andersson:2009}\cite{Wiki:BAC}.

### 2015-10-13 manuscript/content/method.tex
Content fully revised.

@@
 
-The Microsoft Band we used has an assortment of interesting physical and virtual sensors that we can look at, including: an accelerometer, a gyroscope, distance, heart rate, pedometer, skin temperature, and ultraviolet level.
-With the sample size of five volunteers and a controlled setting for the experiments, some will not be as useful for this study: distance, pedometer, and ultraviolet level.
-In fact, their usefulness may be limited even with larger datasets.
+The Microsoft Band we used has an assortment of interesting physical and virtual sensors, including: accelerometer, gyroscope, distance, heart rate, pedometer, skin temperature, and ultraviolet level.
+With our sample size of only five volunteers and a controlled setting for the experiments, some of these sensors will not be very useful for this study, such as the: distance, pedometer, and ultraviolet level, sensors.
+In fact their usefulness may be limited even with larger datasets.
 
-We begin our analysis with the heart rate (HR) by normalizing the heart rates and BACs per subject.
+We begin our analysis with the heart rate (HR) data by normalizing the heart rates and BACs per subject.
 This way we can plot and compare the HR values over the BAC and see if there are any obvious patterns.
@@
 	\includegraphics[width=1.0\textwidth]{../figs/heart_rates}
-	\caption{Heart Rate and BAC per subject (normalized).}
+	\caption{Heart Rate and BAC per subject (normalized)}
 	\label{fig:heart_rate_facet}
@@
 	\includegraphics[width=1.0\textwidth]{../figs/heart_rate_split}
-	\caption{Heart rate split into baseline and excited states.}
+	\caption{Heart rate split into baseline and excited states}
 	\label{fig:heart_rate_split}
@@
 	\includegraphics[width=1.0\textwidth]{../figs/skin_temperatures}
-	\caption{Skin Temperature and BAC per subject (normalized).}
+	\caption{Skin Temperature and BAC per subject (normalized)}
 	\label{fig:skin_temperatures}
@@
 This is because of the well known vasodilation effect of alcohol causing warm blood to come near the surface of the skin \cite{Dekker:1999}.
-It is also known that with a high enough level of intoxication it behaves as a vasoconstrictor and drops skin temperature, which we see with subjects 4 and 5 who had reached the highest BAC of the group.
+It is also known that with a high enough level of intoxication it behaves as a vasoconstrictor and drops skin temperature, which we see with subjects 3 and 4 who had reached the highest BAC of the group.
 Subject 2 had a rapid decrease in temperature at the beginning (we may have started collection too soon before the smartwatch sensor adjusted to his skin temperature), it may be appropriate to discard this portion.
@@
 They might have some hidden patterns that contribute to the performance of a modeling technique.
-We think it's usefulness will be increased if we can use it first to predict whether a user is taking a drink, then provide a time-based aggregate of the estimated number of drinks as a feature in our model to determine drunkenness.
+We think its usefulness will be increased if we can use it first to predict whether a user is taking a drink, then provide a time-based aggregate of the estimated number of drinks as a feature in our model to determine drunkenness.
 

### 2015-10-13 manuscript/conclusions.tex
Restructured paper directory for conference submission.

--- 2015-09-29 manuscript/content/conclusions.tex
@@
 Ideally, we would want a data set from a thousand volunteers in candid situations over several days.
-The two biggest problems are, how do we label the data with the alcohol levels, and how can we model this enormous amount of data in reasonable time?
-Also, will we need to first determine the activity of the user, or will the other sensor provide sufficient information?
+The two biggest problems are, how do we label the data with the alcohol levels in this scenario, and how can we model this enormous amount of data in reasonable time?
+Also, will we need to first determine the activity of the user, or will the other sensors provide sufficient information?
 And if the former, how can we consistently tag these activities in the data?

### 2015-10-13 manuscript/intro.tex
Restructured paper directory for conference submission.

@@
 \section{Introduction}
-%%Describe the problem, state your contributions and the organization of the paper.T
-
 Internet of Things (IoT) is a domain that represents the next most exciting technological revolution since the Internet.
@@
 In the healthcare domain, IoT promises to bring personalized health tracking and monitoring ever closer to the consumers.
-This phenomena is evidenced in a recent Wall Street Journal (June, 29, 2015) article entitled "Staying Connected is Crucial to Staying Healthy".
-Modern smartphones and related devices now contain more sensors than ever before.
-Data from sensors can be collected more easily and more accurately.
-In 2014, it is estimated that 46 million people are using IoT-based health and fitness applications.
-Currently, the predominant IoT-based health applications are in sports and fitness.
-However, disease management or preventive care health applications are becoming more prevalent.
-For example, the real-time preventive care applications such as those for detecting fall in elderly patients is one of the active research areas due to the aging population \cite{Tacconi:2011}.
-This paper describes a new category of real-time preventive health application that aims to reduce risky behaviors due to intoxication by predicting blood alcohol content (BAC) non-invasively and in real-time.
+This phenomena is discussed in a recent Wall Street Journal article, "Why Connected Medicine Is Becoming Vital to Health Care" \cite{Landro:2015}.
+Modern smartphones and smartwatches now contain a more diverse collection of sensors than ever before, and people are warming up to them.
+In January 2014, approximately 46 million US smartphone owners were reported to have used health and fitness applications \cite{Nielsen:2014}.
+Currently, sports and fitness are the predominant foci of IoT-based health applications.
+However, applications in disease management and health care are becoming increasingly prevalent.
+For example, detecting falling of elderly patients \cite{Tacconi:2011}.
 
-The Centers for Disease Control and Prevention (CDC) reported \cite{Schoenborn:2013} that between the years 2008 and 2010 roughly two-thirds of adults were current drinkers, but adults between the ages 18 and 24 were most commonly associated with consumption of more than 5 drinks in one day.
-There is also a known causal relationship between alcohol consumption and risky behaviors \cite{Assaad:2006}.
-One of the worst risky behaviors from drinking is drunk driving which is a persistent problem worldwide.
-Innocent people are harmed/killed in accidents where the driver was under the influence.
-It is reported by the Bureau of Transportation Statistics \cite{Chambers:2012} that in 2010 47.2\% of pedestrian fatalities and 39.9\% of vehicle occupant fatalities are caused by drunk driving.
+Drunk driving is a dangerous, worldwide problem.
+This problem is not only a hazard to the drunk drivers, but also to pedestrians and other drivers.
+It is reported by the Bureau of Transportation Statistics that in 2010, 47.2\% of pedestrian fatalities and 39.9\% of vehicle occupant fatalities were caused by drunk driving \cite{Chambers:2012}.
+The Centers for Disease Control and Prevention (CDC) reported that between the years 2008 and 2010, roughly two-thirds of adults were drinkers, with adults between the ages 18 and 24 having the greatest association with heavy drinking \cite{Schoenborn:2013}.
 
-Being able to calculate BAC is important for the portion of the population that consumes alcohol.
-Drinkers often find themselves in situations in which they are drinking but also need to plan on driving home.
-The issue comes when the drinker is not sure if they are sober enough, as in their BAC is below the legal limit, to drive home safely.
-In many cases, drinkers misjudge their intoxication, believing they will be able to safely get home, when actually their BAC is higher than the legal limit.
-While there are some ways available for people to measure their BAC, these methods are often invasive, inaccurate, or impractical.
+At dangerous levels of intoxication, it can be difficult to judge ones own drunkenness.
+Instead it would be better to get a definitive measurement of the BAC, or simply a binary response: "drunk" or "not drunk." Compact breathalyzers are probably the best option at the moment, but these are not discreet and require deliberate action by the user.
+The other option is to use a smartphone application to manually calculate BAC, but these demand a greater deal of involvement from the user.
+To be practical, it would be useful to have some sort of non-invasive and accurate monitoring system that will warn its user if they become too intoxicated.
+It has been shown that electronic intervention programs are more successful at reducing college student drinking than a general alcohol awareness program \cite{Ward:2015}.
+This system can also be used to warn friends and family, or prevent the operation of the user's car.
 
-The most commonly known method of measuring BAC is the use of a breathalyzer.
-These devices are portable, easily accessible, and accurate.
-However, breathalyzers have a stigma of DUI arrest, fines, and license suspension.
-Breathalyzers are also non-discreet and invasive because the device requires the user to blow into the device in order to measure BAC.
-Due to these reasons, people often do not want to use such a device.
-A noninvasive, discreet, and real-time method of measuring BAC is desired.
-
-In this paper, we propose an android-based IoT middleware architecture and application for collecting sensor data on smartwatches and smartphones to predict BAC in real-time.
-A study conducted in Washington University \cite{Ward:2015} shows that electronic intervention programs are more successful at reducing college student drinking than general alcohol awareness program.
-By wearing a smartwatch, a user can be alerted in real time when the BAC is above certain level and programmed with different options of intervention when such an event occurs.
-For example, the application can send a text message to a trusted family member or friend or can lock the ignition of the car engine of the drinker.
-Currently, there are no known applications that leverages data from common wearable devices to predict BAC non-invasively and in real time.
-
-The main contributions of the paper are:
-\begin{itemize}
-	\item A middleware platform that can collect data from any type of device (virtual or physical) that can communicate with android phones. 
-	\item A BAC prediction model based on live smartwatch sensor data that enables users to get BAC measurement in real time.
-	\item Geographical visualization of BAC predictions.
-	\item Collection of a labeled BAC data set for training the prediction model.
-\end{itemize}
-
-%****Summary of the results and conclusion****
+In this paper, we investigate the prediction of intoxication level from smartwatch sensor data via machine learning.
+We also briefly discuss a general Android-based gateway system which can collect data from any type of physical or virtual sensor accessible by the host smartphone.

### 2015-10-13 manuscript/system.tex
Restructured paper directory for conference submission.

@@
 \section{System Architecture}
-A custom Android application (Figure \ref{app}) was designed to be used during the study.
-This application collected and locally stored sensor data from a Microsoft Band smartwatch in a CSV file.
-The following data was collected from the Microsoft Band:
-\begin{itemize}
-	\item Accelerometer
-	\item Gyroscope
-	\item Pace
-	\item Speed
-	\item Distance traveled
-	\item Motion type (idle, walking, etc.)
-	\item Heart rate
-	\item Pedometer
-	\item Skin temperature
-	\item Ultraviolet level
-	\item Contact (if smartwatch is on)
-	\item Calories burned
-\end{itemize}
-Additional data continuously collected and stored in the CSV by the application included:
-\begin{itemize}
-	\item GPS coordinates from the smartphone 
-	\item Button that was pressed whenever the user took a drink
-	\item Input field used to periodically record the BAC measured by a separate breathalyzer.
-\end{itemize}
 
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=.20]{smartwatch.png}
-	\caption{Capture of the Android application}
-	\label{app}
+A huge end-goal of ours is to have a public system where participating users' phones transmit sensor information to a central database through a REST-based web service.
+The service stores this information in the database and another data processing system.
+Other users who are not contributing information can connect to our system to access other related services, like a choropleth map of the sensor data; BAC prediction data in our case, but the system would not be limited to it.
+The particular protocols and setup used should be designed to protect the privacy of the sensor data contributors.
+
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{figs/system}
+	\caption{Example general system for large-scale sensor data collection}
+	\label{fig:system}
 \end{figure}
 
-User information was also recorded and stored through the application before data collection began.
-Each user was given an unique user id (UUID) and the user profile included age, sex, body mass index, and blood type which was also stored locally as a JSON file.
+\subsection{Smartphone Sensor Gateway}
 
-After the data had been collected, the stored csv file and json file were uploaded to the web server and data inserted into mySQl database by a php script.
-Data stored on the MySQL database was then displayed on a web client by a geographical map.
+Each of the smartphones in this system are behaving as a type of sensor gateway.
+In order to collect our data, we built such a system on the Android platform.
+Though we used primarily a Microsoft Band, our system allows the easy addition of any sensor that can be connected to from the Android smartphone.
+In Figure \ref{fig:gateway}, we show an overview of the classes used in this system.
 
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=.35]{webpage.png}
-	\caption{Current data visualization}
-	\label{webpage}
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{figs/gateway}
+	\caption{Gateway implementation for our data collection application}
+	\label{fig:gateway}
 \end{figure}
-
-As of now, the web client displays the last sample for each user at the recorded location (Figure \ref{webpage}).
-Ideally in future systems, many samples will be averaged across regions (Figure \ref{concept}).
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=.23]{webconcept.png}
-	\caption{Ideal data visualization}
-	\label{concept}
-\end{figure}
-
-The collected data was also used to train a machine learning model which would be implemented in a future Android application to predict user BAC or drunkenness.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=.30]{applicationsystem.png}
-	\caption{Application system}
-	\label{appsystem}
-\end{figure}
-
-Representation of this system can be found in Figure \ref{overallsystem}.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=.14]{overall_system.png}
-	\caption{Overall system}
-	\label{overallsystem}
-\end{figure}
